Hemodynamics regulate spatiotemporal artery muscularization in the developing circle of Willis

Version 1

Abstract

Abstract

Brain arteries are wrapped by vascular smooth muscle cells (VSMCs). Fully differentiated VSMCs are important for brain artery homeostasis, and they are lost in several cerebrovascular diseases. How healthy VSMCs differentiate on different brain arteries during development is unclear. Such knowledge will help regenerate lost VSMCs in brain arteriopathy. To answer this question, we studied the developmental muscularization of the zebrafish circle of Willis (CW) arteries, the major arterial loop that supplies blood to the brain in all vertebrates. We found that artery specification of CW endothelial cells (ECs) happens after they migrate from primitive veins to form CW arteries. VSMCs differentiate from pdgfrb+ common vascular mural cell progenitors at the time when embryo circulation starts and progress temporally and spatially from anterior to posterior CW arteries. Computational fluid dynamic simulation confirms that earlier VSMC differentiation coincide with higher pulsatile flow hemodynamics in anterior CW arteries. Pulsatile blood flow induces the differentiation of human brain pdgfrb+ progenitors into VSMCs and reducing pulsatile blood flow by blocking the zebrafish embryo heartbeat after pdgfrb+ recruitment but before VSMC differentiation limits the number of mature VSMCs. Congruently, the flow responsive transcription factor klf2a is activated in ECs before VSMC differentiation and knockdown delays VSMC differentiation on CW arteries. Overall, our data place hemodynamic activation of endothelial klf2a signaling as key determinant of spatiotemporal VSMC differentiation on CW arteries.

Introduction

Vascular smooth muscle cells (VSMCs) are contractile mural cells wrapping around endothelial cells (ECs) of large vessels, especially arteries (Ando et al., 2022; Donadon & Santoro, 2021; Stratman et al., 2020). The expression of contractile proteins, such as alpha-smooth muscle actin (encoded by acta2), distinguishes VSMC from pericytes, the other type of mural cells primarily associated with small vessels including capillaries (Bahrami & Childs, 2020; Donadon & Santoro, 2021; Stratman et al., 2020). VSMCs are important for arterial homeostasis (Basatemur, Jorgensen, Clarke, Bennett, & Mallat, 2019). In the brain, VSMC constriction and relaxation is essential for functional hyperemia in neurovascular coupling (Hill et al., 2015; Kaplan, Chow, & Gu, 2020). Phenotype switching or dedifferentiation of VSMCs, marked by lower expression of contractile proteins, is prominent in pathological progression of atherosclerosis (Basatemur et al., 2019). Recently, similar VSMC phenotype switching or dedifferentiation have also been described in cell, animal, and human studies of children and adult neurological and cerebrovascular conditions (Aguilar-Pineda et al., 2021; Chou et al., 2022; Milewicz et al., 2010; Oka et al., 2020). These studies suggested that understanding the development of VSMC differentiation on brain arteries might help restore normal VSMC contractility to alleviate various cerebrovascular diseases; such a regenerative strategy requires a deep understanding of how VSMCs acquire a contractile phenotype while differentiating from their progenitors.

VSMCs spatially differentiate from different progenitors from mesoderm and neural crest origin (Ando et al., 2019; Donadon & Santoro, 2021; Whitesell et al., 2019). Previous research described various mechanisms regulating the differentiation of VSMC positive for acta2 in the trunk vasculature. Arterial Notch signaling activated by blood flow is necessary for acta2+ VSMC appearance on the zebrafish dorsal aorta (X. Chen, Gays, Milia, & Santoro, 2017). Autonomous Notch activation is required for specification of pdgfrb+ mural cell progenitors from mesenchyme around arteries, and these progenitors later differentiate into acta2+ VSMCs (Ando et al., 2019). Chemokine signaling promotes VSMCs association with the zebrafish dorsal aorta in the trunk, whereas blood flow modulated transcription factor krüppel-like factor 2 (encoded by klf2a in zebrafish) prevents their association with the adjacent cardinal vein (Stratman et al., 2020). Differentiation into acta2+ VSMCs from pdgfrb+ progenitors in the brain requires autonomous expression of an ATP-sensitive potassium channel (Ando et al., 2022). In contrast, acta2+ VSMCs in the ventral head of zebrafish are not derived from pdgfrb+ progenitors, and their differentiation is regulated by endothelial BMP signaling (Watterston, Zeng, Onabadejo, & Childs, 2019; Whitesell et al., 2019). These studies suggested that VSMC differentiation is highly organotypic and may even be vessel specific.

The circle of Willis (CW) consists of major arteries that supply blood to the vertebrate brain, including the internal carotid arteries and posterior communicating arteries (Campbell et al., 2019; Schröder, Moser, & Huggenberger, 2020). CW arteries are wrapped by VSMCs, and VSMC dedifferentiation and hyperplasia are described in carotid atherosclerosis and the pediatric Moyamoya disease (Chou et al., 2022; Fox, Dorschel, Lawton, & Wanebo, 2021). The internal carotid arteries in the CW are among the earliest circulated arteries in the brain due to their connection to the common carotid arteries (lateral dorsal aorta in zebrafish) (Campbell et al., 2019; Isogai, Horiguchi, & Weinstein, 2001; Schröder et al., 2020). VSMC differentiation on CW arteries, particularly internal carotid arteries, and whether this differentiation is associated with early arterial blood flow in the brain, have not been investigated. The mouse CW resembles that of humans, but its VSMC differentiation is difficult to observe in vivo, as the embryos develop in utero and depend on maternal circulation (Isogai et al., 2001; Schröder et al., 2020). The zebrafish CW also resembles that of humans, and the embryos external growth and optical clarity allows for confocal live imaging of developing blood vessels (Isogai et al., 2001). In addition, a few fluorescent transgenic (Tg) lines have been generated to label VSMCs and their progenitors (Ando et al., 2016; Whitesell et al., 2014). Taking advantage of the zebrafish model, we found a spatiotemporal pattern of VSMC differentiation on CW arteries, and we associated this pattern with hemodynamic signaling pathways in arterial ECs.

Results

Artery muscularization is spatiotemporally regulated in the CW arteries

Arterial ECs in zebrafish brain emerge from primitive veins (Fujita et al., 2011). Recent single-cell transcriptome profiling of vascular cells in prenatal human brain also supported the fate progression from venous to arterial ECs (Crouch et al., 2022). To establish the stage of CW artery specification in the zebrafish brain, we performed confocal live imaging of Tg(flt4:yfp)hu4881, which labels venous ECs, and Tg(kdrl:hras- mcherry)s896, which labels all ECs, at four stages starting at 32 hour post-fertilization (hpf), when CW arteries are assembling (Chi et al., 2008; Hogan et al., 2009; Isogai et al., 2001). As arterial ECs express higher kdrl, we used the intensity of mcherry as an indicator of artery specification (Figure 1A-1E) (Chi et al., 2008). We found that ECs in CW arteries are primarily venous at 32 hpf (Figure 1A and 1E), and gradually gain kdrl expression from 54 hpf to 3-day post fertilization (dpf) (Figure 1B, 1C, and 1E). The mcherry intensity at 4 dpf is similar to 3 dpf, suggesting that artery specification of the CW is completed by 3 dpf (Figure 1C-1E).

Artery specification of endothelial cells (ECs) in circle of Willis (CW) arteries

(A-D) Confocal live images of Tg(flt4:yfp, kdrl:hras-mcherry)hu4881/s896 and scheme representation of CW arteries in zebrafish brain at 32 hour post fertilization (hpf) (A), 54 hpf (B), 3 day post fertilization (dpf) (C), and 4 dpf (D). Green channel represents flt4:yfp fluorescence, Red channel represents kdrl:hras-mcherry fluorescence, and Merge panel combines both channels. Arrows point to the CW arteries with kdrl:hras-mcherry signal. Scale bar = 50 μm

(E) Average intensity of kdrl:hras-mcherry in caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=6, 2 independent experiments), 54 hpf (n=12, 4 independent experiments), 3 dpf (n=15, 4 independent experiments), and 4 dpf (n=9, 2 independent experiments), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, * p≤0.05, ** p≤0.01

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Then we characterized CW artery muscularization. Previous research on the CW arteries described that VSMCs derive from pdgfrb+ mural cell progenitors (Ando et al., 2022; Ando et al., 2019). Thus, to determine VSMC differentiation in real time we performed confocal live imaging of Tg(acta2:mcherry)ca8, TgBAC(pdgfrb:egfp)ncv22, and Tg(kdrl:cerulean)sd24, to visualize respectively VSMCs (red), pdgfrb+ progenitors (green) and the developing arteries (white) at four stages starting from 32 hpf (Ando et al., 2016; Page et al., 2013; Whitesell et al., 2014). We found that overall pdgfrb+ progenitors appear on CW arteries after their lumen form around 54 hpf, and most of them do not express acta2 (Figure 2A-2B, 2E-2F, and S1A-S1B). The number of pdgfrb+ progenitors increase temporally from 54 hpf to 3 dpf and spatially on CW arteries from anterior to posterior - from the internal carotid arteries (CaDI), the basal communicating artery (BCA) to the posterior communicating segments (PCS) (Figure 2B-2C). Notably, while many pdgfrb+ progenitors on CaDI differentiate into acta2+ VSMCs by 3 dpf nearly no acta2+ VSMCs were observed in the nearby artery BCA or PCS (Figure 2B-2C, 2E-2F, and S1A-S1B). By 4 dpf, the number of pdgfrb+ cells is similar to 3 dpf, but acta2+ VSMCs on CaDI continue to differentiate in a greater number over the BCA, and PCS arteries (Figure 2C-2F and S1A-S1B). These results show an anterior to posterior differentiation of pdgfrb+ progenitors into acta2+ VSMCs with the CaDI being the first CW artery muscularized and having a higher number of acta2+ VSMCs.

Vascular smooth muscle cell (VSMC) differentiation on circle of Willis (CW) arteries

(A-D) Confocal live images of Tg(acta2:mcherry, kdrl:cerulean)ca8/sd24 TgBAC(pdgfrb:egfp)ncv22 and scheme representation of vascular endothelium and mural cells on CW arteries in zebrafish brain at 32 hour post fertilization (hpf) (A), 54 hpf (B), 3 day post fertilization (dpf) (C), and 4 dpf (D). White channel represents kdrl:cerulean fluorescence, Red channel represents acta2:mcherry, Green channel represents pdgfrb:egfp, Merge 1 panel combines all three channels, Merge 2 combines acta2:mcherry in red and pdgfrb:egfp in green. Arrows point to the CW arteries with pdgfrb:egfp and acta2:mcherry signal. Scale bar = 50 μm

(E) Number of pdgfrb+ vascular mural cell progenitors per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=3, 1 independent experiment), 54 hpf (n=6, 1 independent experiment), 3 dpf (n=6, 1 independent experiment), and 4 dpf (n=5, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, ** p≤0.01, *** p≤0.001, **** p≤0.0001

(F) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 32 hpf (n=5, 1 independent experiment), 54 hpf (n=6, 2 independent experiments), 3 dpf (n=24, 6 independent experiments), and 4 dpf (n=25, 6 independent experiments), two- way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

CW arteries have spatiotemporal difference in hemodynamics

Our data suggest that VSMCs in the anterior CaDI differentiate significantly earlier than the anterior BCA and the posterior PCS, even if their progenitors have the same origin and these CW blood vessel formation and arterialization occurs at the same time. We speculate that different hemodynamic distribution may contribute to the spatiotemporal difference in VSMC differentiation on CW arteries. In fact, CaDI is the only vessels in the CW receiving pulsatile arterial blood flow directly from the lateral dorsal aorta and cardiac outflow tract. Furthermore, the CaDI differs from the BCA or PCS in terms of overall vascular geometry, which has been shown to affect local intravascular forces such as the wall shear stress (WSS) (Katritsis et al., 2007). To explore this possibility, we performed microangiography with a high molecular weight fluorescent dextran and used computational fluid dynamics (CFD) to simulate the effect of pulsatile flow on the CW intravascular forces (ANSYS, 2014; Z. Chen et al., 2019; Fernandes et al., 2022). First, we determined each CW artery diameter, the principal determinant of hemodynamics, and found that there was no significant difference between structures (Figure 3A-3D). Notably, however, the flow velocity across CW arteries was differentially distributed, with the CaDI having the highest blood flow speed starting from 54 hpf to 4 dpf (Figure 3A-3C, and 3E). This is consistent with the CaDI being the inlet vessel for the arterial blood flow within the entire CW circulation. Next, we calculated the WSS across all CW arteries and found an overall increase in average and total WSS from 54 hpf to 4 dpf (Figure 3A-3C, and 3F). Importantly, however, when analyzed individual vessels, we found that the WSS in the CaDI is significantly higher from 54 hpf to 4 dpf, while this increase was not noted in the BCA or PCS (Figure 3A-C, and 3G). Further, WSS in the CaDI is significantly higher than in the BCA and PCS at 3 dpf (Figure 3G), when considerable VSMC differentiation starts on the CaDI but not the BCA or PCS (Figure 2B-2C, 2F, and S1A-S1B). Altogether, the pulsatile blood flow simulation suggests that CW arteries experience different hemodynamic loads that might confer an heterogenous muscularization of CW arteries.

Computational fluid dynamic (CFD) simulation of circle of Willis (CW) arteries

(A-C) Confocal live images of Tg(kdrl:gfp)zn1 injected with dextran and representation of vascular diameter, CFD simulated wall shear stress (WSS), and flow velocities in CW arteries at 54 hour post fertilization (hpf) (A), 3 day post fertilization (dpf) (B), and 4 dpf

(C). Arrows point to the CW arteries with high flow velocity and WSS. Scale bar = 50 μm

(D) Average vascular diameter in caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 54 hpf (n=6, 1 independent experiment), 3 dpf (n=6, 1 independent experiment), and 4 dpf (n=6, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, * p≤0.05, ** p≤0.01

(E) Average flow velocity in CaDI, BCA, and PCS at 54 hpf (n=3, 1 independent experiment), 3 dpf (n=3, 1 independent experiment), and 4 dpf (n=3, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, * p≤0.05, ** p≤0.01

(F) Average wall shear stress (WSS) throughout the CW arteries at 54 hpf (n=3, 1 independent experiment), 3 dpf (n=3, 1 independent experiment), and 4 dpf (n=3, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, **** p≤0.0001

(G) Average wall shear stress (WSS) in CaDI, BCA, and PCS at 54 hpf (n=3, 1 independent experiment), 3 dpf (n=3, 1 independent experiment), and 4 dpf (n=3, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, * p≤0.05, ** p≤0.01

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, WSS; wall shear stress, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Blood flow is required for differential CW artery muscularization

Our data suggest that ECs under pulsatile arterial flow might favor the differentiation of acta2+ VSMCs from pdgfrb+ progenitors. To test this hypothesis, we set up an in vitro cell co-culture experiment where GFP-PDGFRB+ human brain pericytes (GFP-HBVPs) were cultured in flow amenable outlet slides, then covered by a thin layer of collagen type 1. Confluent human endothelial cells (HUVECs) were seeded on top of that and then exposed to steady-state laminar flow or pulsatile flow conditions (Figure 4A) (Abello, Raghavan, Yien, & Stratman, 2022). Maturation of PDGFRB+ human brain pericytes into ACTA2+ VSMCs was then quantified after 24 hours from the introduction of flow addition. We found that brain pericytes under pulsatile flow were larger and showed higher expression of VSMC differentiation markers including PDGFRB, ACTA2, and TRANSGELIN (Robin et al., 2013) (Figure 4B-4D), suggesting that pulsatile blood flow favored PDGFRB+ progenitor differentiation.

Blood flow is required for vascular smooth muscle cell (VSMC) differentiation on circle of Willis (CW) arteries

(A) Scheme representation of in vitro cell co-culture experiment

(B) Representative immunofluorescence images of brain pericytes after exposure of pulsative flow and laminar flow. Cells were stained for ACTA2 (cyan), and cytosolic GFP label (magenta). Scale bar = 10 μm

(C) Morphological measurement of brain pericyte length/width ratio before and after exposure of pulsative flow and laminar flow, 3 independent experiments, two-tailed Mann–Whitney test, represented with mean ± SD, **** p≤0.0001

(D) Protein level of PDGFRB (3 independent experiment), ACTA2 (5 independent experiment), and TAGLN (TRANSGELIN, 3 independent experiment) in arbitrary unit (A.U.) after exposure of pulsative flow and laminar flow, two-tailed Mann–Whitney test, represented with mean ± SD, **** p≤0.0001

(E-G) Confocal live images of Tg(acta2:mcherry; kdrl:gfp)ca8/zn1 and scheme representation of vascular endothelium and VSMCs on CW arteries at 3 day post fertilization (dpf) in control embryos (E), embryos injected with 0.35 ng tnnt2a morpholino (MO) (F), embryos treated with 25 μM nifedipine from 54 hour post fertilization (hpf) (G). Red channel represents acta2:mcherry, Green channel represents kdrl:gfp, and Merge panel combines both channels. Arrows point to the CW arteries with acta2:mcherry signal. Scale bar = 50 μm

(H) Number of acta2+ VSMCs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 3 dpf in uninjected control (n=6, 1 independent experiment) and embryos injected with 0.35 ng tnnt2a MO at one to two cell stage (n=6, 1 independent experiment), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, *** p≤0.001, **** p≤0.0001

(I) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in uninjected control (n=6, 1 independent experiment) and embryos injected with 0.35 ng tnnt2a MO at one to two cell stage (n=6, 1 independent experiment), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, **** p≤0.0001 (J-L) Confocal live images of Tg(acta2:mcherry; kdrl:gfp)ca8/zn1 and scheme representation of vascular endothelium and VSMCs on CW arteries at 4 dpf in control embryos (J), embryos injected with 0.35 ng tnnt2a MO (K), embryos treated with 25 μM nifedipine from 54 hpf (L). Red channel represents acta2:mcherry, Green channel represents kdrl:gfp, and Merge panel combines both channels. Arrows point to the CW arteries with acta2:mcherry signal. Scale bar = 50 μm

(M) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 3 dpf in DMSO control (n=8, 2 independent experiments) and embryos treated with 25 μM nifedipine from 54 hpf (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01, **** p≤0.0001

(N) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in DMSO control (n=9, 2 independent experiments) and embryos treated with 25 μM nifedipine from 54 hpf (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, MO: morpholino, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Next, we tested the effect of pulsatile flow in the CW on VSMC differentiation in vivo. Previous research suggested that blood flow hemodynamics is not necessary for pdgfrb+ progenitor recruitment at CW arteries, as shown by zebrafish embryos lacking blood flow such as the tnnt2a morphants (MO) (Ando et al., 2016), however their differentiation to acta2+ cells has not been documented. Thus, to determine whether blood flow affects acta2+ VSMC differentiation on CW arteries, we injected 0.35 ng of tnnt2a MO into Tg(acta2:mcherry, kdrl:gfp)ca8/zn1 at the one to two-cell stage, to abrogate cardiac contractility and thus pulsatile flow (Sehnert et al., 2002; Whitesell et al., 2014; Zhong et al., 2006). Notably, we found no acta2+ VSMCs on CW arteries of tnnt2a MO at 3 to 4 dpf (Figure 4F, 4H-4I, and 4K), suggesting that blood flow is dispensable for pdgfrb+ progenitor recruitment but is required for acta2+ VSMC differentiation on the CW arteries.

To further explore the temporal requirement of blood flow, we treated embryos with 25 μM nifedipine, a drug shown to reduce heart rate in zebrafish embryos (Gierten et al., 2020), from 54 hpf, right before the onset of pdgfrb+ progenitors’ expression of acta2 (Figure. 2A and B). We found that after 18 hours at 3 dpf the number of acta2+ VSMCs is greatly reduced on the CaDI of treated embryos compared to untreated controls (Figure 4G and 4M); acta2+ VSMCs on the BCA and the PCS at 4 dpf are also reduced significantly (Figure 4L and 4N), suggesting that blood flow reduction after 54 hpf delays VSMC differentiation on CW arteries with the earliest and strongest effect on the CaDI muscularization (Figure 4G and 4M).

To determine whether blood flow is differentially required for the maintenance of VSMCs on CW arteries, we treated embryos with nifedipine from 4 dpf, after VSMC differentiation on CW arteries, and imaged at 5 dpf. The number of acta2+ VSMCs on the CaDI and PCS in treated embryos is similar to untreated controls (Figure S2A-S2C), suggesting that blood flow is required for differentiation but not for short-term maintenance of VSMCs.

Blood flow-regulated transcription factor klf2a is required for spatiotemporal CW artery muscularization

Our results suggest that ECs might express a flow pattern-dependent transcriptional program that favors spatiotemporal VSMC muscularization in the CW arteries. For example, previous studies showed that the site and level of the transcription factor Klf2 (Parmar et al., 2006; Sweet, Fan, Hsieh, & Jain, 2018; Warboys, Amini, de Luca, & Evans, 2011) in artery ECs closely follow the predicted pattern of elevated intravascular forces (Lee et al., 2006). Consistently, higher Klf2 expression can be induced by unidirectional pulsatile flow mimicking arterial flow (Dekker et al., 2002). Furthermore, klf2 is implicated in VSMC migration on arterial blood vessels during mouse development: loss of klf2 leads to aorta VSMC deficiency (Wu, Bohanan, Neumann, & Lingrel, 2008) while endothelial klf2a expression prevents VSMC association with primitive veins in zebrafish trunk (Stratman et al., 2020). Hence, we tested the hypothesis that klf2a might be the signaling spatially regulated in the CW artery to control VSMC maturation. We first imaged Tg(klf2a:h2b-egfp, kdrl:hras- mcherry)ig11/s896, which labels nuclei with active klf2a expression, and quantified the number of klf2a+ ECs in the CW arteries (Chi et al., 2008; Steed et al., 2016). To account for the overall changes in EC number during development (Ulrich, Ma, Baker, & Torres-Vazquez, 2011), as well as the length of each vessels, we also imaged Tg(fli1:nls-gfp, kdrl:hras-mcherry)y7/s896, which labels all EC nuclei, at four stages from 32 hpf (Roman et al., 2002). Notably, we found a slight increase in ECs number per 100 μm vessel in the CaDI from 32 to 54 hpf (Figure S3A-S3C), but a much more significant increase in klf2a+ EC nuclei in the CaDI (Figure 5A and 5B), before acta2+ VSMCs appear (Figure 2B-2C, 2F, and Figure S1A-S1C). The number of ECs in the CaDI stayed the same from 54 hpf through 4 dpf (Figure S3A-S3C), while klf2a+ ECs in the CaDI increased until 3 dpf (Figure 5A and 5B). In tandem, acta2+ VSMCs on the CaDI increase through 4 dpf (Figure 2C-2D, 2F, and Figure S1A-S1C). EC number per 100 μm vessel in the BCA and PCS are similar from 54 hpf to 4 dpf (Figure S3A-S3C), whereas the number of klf2a+ ECs increase significantly over time (Figure 5A and 5B). This increase predates the presence of acta2+ VSMCs on the BCA and PCS (Figure 2B-2D, 2F, and Figure S1A-S1C). These results suggest that the increase of klf2a+ ECs in the CW arteries is not due to an increase in EC number and precedes the spatiotemporal VSMC differentiation observed on the CW arteries.

Blood flow regulated transcription factor klf2a is expressed in circle of Willis (CW) arteries

(A) Confocal live images of Tg(klf2a:h2b-gfp, kdrl:ras-mcherry)ig11/s896 and scheme representation of endothelial cells (ECs) in CW arteries in zebrafish brain at 32 hour post fertilization (hpf), 54 hpf, 3 day post fertilization (dpf), and 4 dpf. Green channel represents klf2a:h2b-gfp, Red channel represents kdrl:ras-mcherry, and Merge panel combines both channels. Arrows point to the CW arteries with klf2a:h2b-gfp signal. Scale bar = 50 μm

(B) Number of klf2a+ ECs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=6, 3 independent experiments), 54 hpf (n=6, 2 independent experiments), 3 dpf (n=11, 3 independent experiments), and 4 dpf (n=12, 3 independent experiments), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, ** p≤0.01, *** p≤0.001, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

To further define the role of klf2a in CW artery muscularization, we knocked down klf2a in Tg(acta2:mcherry, kdrl:gfp)ca8/zn1 with 11 ng MO injection at the one to two-cell stage and imaged at 3 to 4 dpf (Nicoli et al., 2010; Whitesell et al., 2014; Zhong et al., 2006). We validated the effect of klf2a MO knockdown using the Tg(klf2a:h2b- egfp, kdrl:hras-mcherry)ig11/s896 transgenic line (Figure S3D-S3F) (Chi et al., 2008; Steed et al., 2016). Compared to uninjected control, klf2a morphants have significantly less acta2+ VSMCs on the CaDI at 3 dpf but a normal number by 4 dpf (Figure 6A-6D), suggesting that klf2a promotes timely initiation of CaDI muscularization. Together these data suggest that CW muscularization is associated with endothelial klf2a activation and necessary for spatiotemporal VSMC differentiation under pulsatile blood flow conditions.

klf2a promotes vascular smooth muscle cell (VSMC) differentiation on anterior circle of Willis (CW) arteries

(A-B) Confocal live images of Tg(acta2:mcherry; kdrl:gfp)ca8/zn1 and scheme representation of vascular endothelium and VSMCs on CW arteries at 3 dpf (A) and 4 dpf (B) in uninjected control embryos, embryos injected with 11 ng klf2a morpholino (MO). Red channel represents acta2:mcherry, Green channel represents kdrl:gfp, and Merge panel combines both channels. Arrows point to the CW arteries with acta2:mcherry signal. Scale bar = 50 μm

(C) Number of acta2+ VSMCs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 3 dpf in uninjected control (n=8, 2 independent experiments) and embryos injected with 11 ng klf2a MO at one to two cell stage (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01

(D) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in uninjected control (n=8, 3 independent experiments) and embryos injected with 11 ng klf2a MO at one to two cell stage (n=8, 3 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Discussion

Here, we used confocal live imaging of fluorescence transgenic zebrafish embryos to characterize the spatiotemporal dynamics of VSMC differentiation on the CW, which consists of major arteries that supply blood to the vertebrate brain. We found that CW morphogenesis preceded arterial specification. pdgfrb+ mural cell progenitors start to express the VSMC marker acta2 after these mural cell progenitors were recruited to CW arteries. VSMCs differentiated earlier on anterior CW arteries, which are under higher WSS than their posterior counterparts, due to the high velocity of incoming pulsatile blood flow. We used an in vitro co-culture assay, genetic manipulation, and drug treatments to provide evidence that pulsatile blood flow can contribute to spatiotemporal VSMC differentiation. We found that the flow responsive transcription factor klf2a is activated from anterior to posterior in the CW arteries, preceding VSMC differentiation. klf2a knockdown delayed VSMC differentiation on anterior CW arteries. Together, these data support the conclusion that pulsatile flow activation of endothelial klf2a promotes spatiotemporal VSMC differentiation on CW arteries in the brain (Figure 7).

Schematic model of the developmental muscularization of the CW arteries

The model shows how pulsatile flow generate higher hemodynamics in anterior CW arteries like the caudal division of internal carotid artery (CaDI) via the activation of endothelial klf2a signaling. Other posterior CW arteries with straight shape like the posterior communicating segment (PCS) experience less hemodynamic force and showed moderate klf2a activation and VSMC differentiation.

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, CaDI: caudal division of internal carotid artery, PCS: posterior communicating segment

Our work suggests klf2a-mediated blood flow regulation of VSMC differentiation on zebrafish brain arteries, and thus raises the question of how Klf2 transduces endothelial signals to mural cell progenitors and VSMCs. In vitro, endothelial Klf2 upregulates miR- 143/145, which are transported into co-cultured VSMCs within extracellular vesicles (EV) to promote a contractile phenotype in VSMCs (Hergenreider et al., 2012). How EV transport of miR-143/145 works in vivo during development remains unknown.

Notch signaling appears a plausible downstream effector of Klf2 activation. Like Klf2, Notch also responds to flow in heart valve development (Fontana et al., 2020), and Klf2 appears to upregulate Notch (Duchemin, Vignes, & Vermot, 2019). Non-canonical Notch signaling is induced by shear stress to mediate endothelial barrier formation (Polacheck et al., 2017). Notch1 is a proposed mechanosensor in adult arteries (J. J. Mack et al., 2017). In addition, Notch regulation of VSMC development is well established. Additionally, flow activation of Notch in the dorsal aorta (DA) is required for VSMC recruitment (X. Chen et al., 2017). Notch signaling is also activated in DA VSMC progenitors during development, when DA ECs express Jag1 (jagged canonical Notch ligand 1) (Chang et al., 2012; High et al., 2008). Notch2 and Notch3 are compensatory for VSMC development on the DA (Q. Wang, Zhao, Kennard, & Lilly, 2012). In VSMC differentiation on coronary arteries, ECs express Jag1 and pericyte progenitors express Notch3 (Volz et al., 2015). Thus, reciprocal Notch signaling between ECs and VSMCs may regulate VSMC differentiation on brain arteries, although how Notch signaling activates Myocd (myocardin) (Huang et al., 2008; Li, Wang, Wang, Richardson, & Olson, 2003), Srf (serum response factor) (C. P. Mack & Owens, 1999), and miR- 143/145 (Boettger et al., 2009) to enable expression of Acta2 and other contractile proteins remain incompletely understood.

It is important to recognize the versatile roles of Notch signaling in vascular development for process both prior to and in tandem with VSMC differentiation. Notch signaling represses proliferation of arterial ECs during angiogenesis (Hasan et al., 2017; Pitulescu et al., 2017), and maintains arterial identity after morphogenesis (Lawson et al., 2001; Shutter et al., 2000; Villa et al., 2001). Notch signaling is required for recruitment of mural cells to brain arteries, in which Notch2 and Notch3 are redundant for upregulation of Pdgfrb in mesenchymal progenitors (Ando et al., 2019). Notch3, in which mutations cause cerebral autosomal dominant arteriopathy with subcortical infarcts and leukoencephalopathy (CADASIL), is broadly required for mural cell development in the brain (Domenga et al., 2004; Joutel et al., 1996; Y. Wang, Pan, Moens, & Appel, 2014). Thus, thorough dissection of how Notch signaling regulates VSMC differentiation on brain arteries would require development of novel methods that enable spatiotemporal resolution of Notch ligands and receptors in different vascular cell types.

Wnt signaling is another possible downstream effector of Klf2. In heart valve development, endocardial Klf2 upregulates Wnt9 (wingless-type MMTV integration type family member 9), which activates Wnt signaling in neighboring mesenchymal cells and thereby regulates their proliferation and condensation (Goddard et al., 2017). The pathway interacts with Notch signaling in heart valve morphogenesis (Paolini, Fontana, Pham, Rodel, & Abdelilah-Seyfried, 2021). Non-canonical Wnt signaling attenuates EC sensitivity to shear stress and stabilizes blood vessels (Franco et al., 2016). In brain vascular development, Wnt signaling controls brain specific developmental angiogenesis and endothelial barrier formation (Daneman et al., 2009; Liebner et al., 2008; Zhou et al., 2014). Wnt signaling in mural cells activates expression of Lama2 (laminin subunit alpha 2), a major component of the brain vascular basement membrane, to promote the neurovascular unit and blood-brain barrier maturation (Biswas et al., 2022). Wnt signaling regulates VSMC proliferation, migration, and survival in cardiovascular diseases (Mill & George, 2012), but relatively little is known about the role of Wnt signaling in VSMC differentiation, especially organotypic VSMC differentiation on brain arteries.

The biochemical pathway by which Klf2 activates Notch or Wnt signaling remains incompletely understood. Recent chromatin occupancy and transcription studies showed that Klf2 has context specific binding patterns and transcriptional targets in the heart and lung endothelium (Sweet et al., 2023). Thus, Klf2 may have multiple direct or indirect paths to activate Notch or Wnt signaling. The context specificity may also explain different effects of klf2a on VSMC development in the zebrafish brain and trunk: previous research suggested that klf2a expression in the cardinal vein prevents VSMC association (Stratman et al., 2020), whereas this study suggests that klf2a activation in brain arteries promotes VSMC differentiation. It is plausible that klf2a has different expression levels, binding patterns, transcriptional targets, and downstream effects in brain arterial ECs compared with trunk venous ECs.

Our findings raise interesting new questions on whether stable klf2a expression in CW arterial ECs supports further maturation of VSMCs. Previous research found that acta2+ VSMCs on the BCA and PCS express pericyte enriched abcc9 (ATP binding cassette subfamily C member 9) at 4 dpf (Ando et al., 2022), which is gradually lost from 5 dpf to 6 dpf (Ando et al., 2022) as these cells become acta2+. Interestingly, acta2+ VSMCs on CaDI, BCA, and PCS still retain expression of pdgfrb at 6 dpf (Ando et al., 2021). Thus, it is possible that stable expression of klf2a in CW arterial ECs will support later on complete VSMC maturation, for example in Transgelin positive cells (Colijn, Nambara, & Stratman, 2023). Alternatively, acta2+ VSMCs with pdgfrb expression are maintained as such in the CW arteries (Ando et al., 2021).

It is important to note the differences between the trunk and our data focused on the CW brain arterial VSMC differentiation. On the dorsal aorta of the trunk, as soon as VSMCs are recruited, they begin expressing Acta2 and Tagln, suggesting simultaneous recruitment and differentiation from sclerotome progenitors (Ando et al., 2016; Stratman et al., 2017), compared with the spatiotemporal dynamics we see on brain arteries. It is also important to note the difference between arterial and venous VSMCs. Brain arterial VSMCs strongly express Acta2, whose expression is weak in venous VSMCs (Crouch, Joseph, Marsan, & Huang, 2023; Hill et al., 2015). Venous VSMCs share more common gene expression signature with pericytes, as they both express Abcc9 and Kcnj8 (Ando et al., 2019; Bondjers et al., 2006; Chasseigneaux et al., 2018; He et al., 2016; Vanlandewijck et al., 2018). Based on our data, it is possible to speculate that pulsatile flow reinforces brain arterial specification before VSMC differentiation. Pulsatile flow contributes to both arterial specification and VSMC differentiation, although the two processes are difficult to dissect in vivo, as both require Notch signaling and overlap temporally. In addition, pulsations dampen from dorsal aorta to internal carotid arteries, and flow eventually becomes stable in capillaries and veins (Humphrey & Schwartz, 2021). Current in vivo methods involving genetic manipulations and drug treatments are only capable of qualitative reduction in all types of flow, including pulsatile. In vitro three-dimensional (3D) vascular culture models, which combine ECs and mural cells (Mirabella et al., 2017; Vila Cuenca et al., 2021), could be further optimized to simulate complex geometry of brain arteries. Combining these models with microfluidics, which allows precise calibration of flow velocity and pulse, would enable a more thorough analysis of endothelial mechanotransduction and its contribution to VSMC differentiation on different arterial beds (Abello, Raghavan, Yien, & Stratman, 2022; Gray & Stroka, 2017; Griffith et al., 2020).

In conclusion, our work highlights organotypic differences in VSMC differentiation in the CW brain arteries and points to hemodynamics as a main driver of spatiotemporal dynamics of VSMC maturation on the brain arteries. Impaired VSMC differentiation in development could be implicated in cerebrovascular diseases, such as Moyamoya disease, which is linked to mutations in the human ACTA2 gene that results in VSMC hyperplasia specifically on the internal carotid arteries (Fox et al., 2021; Guo et al., 2007; Guo et al., 2009; Lin et al., 2012). Our study points to endothelial cell mechanotransduction of pulsatile flow as a key signaling network that could render specific VSMC populations on brain arteries more susceptible compared to VSMCs associated with other vessels in the same tissue. Our data will therefore inform the screening of new genes that in combination to known genetic variants, such as ACTA2, could contribute to our understanding of the susceptibilities in cerebrovascular diseases.

Material and Methods

Zebrafish husbandry and transgenic lines Zebrafish were raised and maintained at 28.5 °C using standard methods. Protocols are approved by the Yale Institutional Animal Care and Use Committee (2020-11473). Transgenic lines in Table 1 were established previously.

Confocal fluorescence microscopy

Zebrafish embryos were raised in 0.003% 1-phenyl-2-thiourea (PTU, phenylthiocarbamide, or n-phenylthiourea, Sigma P7629) from gastrulation stage to prevent pigmentation. Embryos imaged live by confocal fluorescence microscopy were anesthetized in 0.1% tricaine methanesulfonate (TMS, MS-222, or Syncaine, Western Chemical, NC0872873) and mounted in 1% low melt agarose within glass bottom microwell dishes. Fluorescence images were captured with an upright Zeiss LSM 980 confocal microscope using a 20X objective.

Image analysis

Confocal fluorescence images were analyzed with Imaris microscopy image analysis software (Bitplane, Oxford Instruments). Average fluorescence intensity of mcherry in Tg(flt4:yfp, kdrl:hras-mcherry)hu4881/s896 was estimated by generating volume objects covering the artery of interest with Surfaces module. Vessel lengths of the arteries were manually traced with Filament module. Numbers of pdgfrb+ mural cell progenitors, acta2+ vascular smooth muscle cells, and klf2a+ and fli1+ endothelial nucleus were counted with Spots module.

Morpholino injections

Morpholino antisense oligonucleotides were synthesized by Gene Tools. Morpholinos in Table 2 were validated previously. Optimized dose injected into each embryo are listed. Uninjected siblings were used as controls.

Nifedipine treatment

Nifedipine was dissolved into 20 mM in dimethyl sulfoxide (DMSO). The stock solution was diluted into 25 μM in egg water with 0.003% PTU to treat zebrafish embryos from 54 hpf to 3 dpf or 4 dpf. The stock solution was diluted into 20 μM in egg water with 0.003% PTU to treat zebrafish embryos from 4 dpf to 5 dpf. The same volume of DMSO was added into egg water with 0.003% PTU for sibling control embryos.

Microangiography

Embryos were anesthetized in 0.1% tricaine methanesulfonate, placed on an agarose mold, and injected pericardially with 4 nL Tetramethylrhodamine Dextran (2,000,000 molecular weight, Thermo Fisher) at a concentration of 10 mg/mL. Subsequently, the vasculature of embryos was checked for Dextran fluorescence signal under a stereomicroscope (Olympus, MVX10). Embryos with Dextran fluorescence were mounted in 1% low melt agarose within glass bottom microwell dishes and imaged with confocal microscope.

Computational fluid dynamic simulation

Computational fluid dynamic simulation was performed as previously described with modification (ANSYS, 2014; Barak et al., 2021). Three-dimensional (3D) geometry models of the circle of Willis (CW) arteries were reconstructed from confocal microangiography images of Dextran injected Tg(kdrl:gfp) embryos with the Filament module in Imaris (Bitplane, Oxford instruments). The CW 3D geometry models were pre-processed in ANSYS SpaceClaim 2022 R1 software. Pre-processed models were then meshed in ANSYS Fluent 2022 R1 software using computational fluid dynamics application settings. Inlets and outlets were specified. Meshing orthogonal quality was calculated, and objects less than 0.01 were excluded. Blood density and gauge pressure were considered as 1,060 kg/m3 and 13,332 pascals, respectively. Each analysis consisted of 200 iterations.

Cell Culture

Pooled primary Human Umbilical Vein Endothelial Cells (HUVEC) (PCS-100-010TM, ATCC) were seeded at an initial concertation of 5,000 cells per cm2, in 1x M119 media (110343-023, Gibco) supplemented with 16% FBS (Gibco), 84 µg/mL of heparin sodium salt (H3393, Sigma-Aldrich), 25 µg/mL of endothelial cell growth supplement (02-102, EMD Millipore Corp.) and 1x Antibiotic-antimycotic solution (15240-062, Gibco). Cell cultures were maintained at 37°C, 5% CO2 and 95% humidity, until the cell reached 80% confluence.

To facilitate cell visualization in cell co-cultures, Human Brain Vascular Pericytes (HBVP) (1200, ScienCell) were transfected using lentiviral particles (LVP310, GenTarget, Inc.) to induce GFP expression under EF1a promotor. Cells cultures were initiated by seeding 5,000 cells per cm2 in 175 cm2 plastic flasks pre-coated with gelatin and 1x DMEM (11995-065, Gibco) supplemented with 10% FBS (Gibco), and 1x Antibiotic-antimycotic solution under the same cell culture conditions described above. When cells cultures were at 80% confluence, 200 µL containing 2 x 106 GFP-lentiviral particles were added to each 175 flasks. After 72 hours, fresh cell culture media supplemented with 10µg/mL of Blasticidin (15205, Sigma-Aldrich) were added to each flask to select the positive transfected cells.

Flow assays and immunostaining

HBVP cells were harvested and seeded at a concentration of 1.3 x 105 cells in 0.4 optical plastic flow microslides (80176, Ibidi) precoated with 1 mg/mL gelatin and incubated for 24 hours under standard culture conditions. After the initial incubation, 100 µg/mL of collagen I (354249, Corning) diluted in DMEM cell culture media was added to the slides to create a thin layer on top of the HBVP cells. After 2 hours, the media was removed and 2.5 x 105 HUVECs were seeded on top of the collagen I layer and incubated for additional 24 hours. After cell co-cultures were established, the slides were exposed to laminar (15 dyn/cm2) or pulsatile (12-15 dyn/cm2) flow for 24 hours, implementing a peristaltic pump adapted to produce different types of flow (Abello, Raghavan, Yien, & Stratman, 2022). After 24 hours, cultures were rinsed with 1x PBS and fixed for 30 minutes in 4% paraformaldehyde at room temperature. Cell cultures were immunostained with α-Smooth muscle actin D4K9N XP® rabbit monoclonal antibody (19245, Cell Signaling), followed by Alexa Fluor 633 goat anti-rabbit IgG (A21071, Invitrogen). Confocal images were obtained using a 40x objective with a W1 Spinning Disk confocal microscope, a Fusion camera, and the Nikon Eclipse Ti2-E base. Fiji image processing software was used for image analysis and fluorescence intensity quantification.

Statistics

All statistical analyses were performed with GraphPad Prism (version 10.0.3). Mann–Whitney test was used to compare two groups to test mean differences (protein level, morpholino and nifedipine treatment). Two-way analyses of variance followed by Tukey’s multiple comparisons was used to compare more than two groups to test mean differences (average fluorescence intensity, cell number, vascular diameter, flow velocity, and WSS).

eLife Assessment

This study showing how the Circle of Willis acquires smooth muscle coverage during development in the zebrafish model is important, and the evidence provided is solid, with only minor weaknesses. The work is of interest to researchers working on cerebral circulation, angiogenesis, and developmental vascular stabilization.

Reviewer #1 (Public Review):

Cheng et al investigated how vascular cells of the zebrafish Circle of Willis arteries differentiate using live imaging of transgenic zebrafish embryos. They find an anterior-to-posterior gradient in the differentiation of pdgfrb+ progenitors into acta2+ smooth muscle cells (SMCs). Computational modeling suggests that blood flow velocity and wall shear stress are higher in the anterior Circle of Willis arteries. Using pharmacological manipulations, they show that blood flow is required for the differentiation of SMCs but not for the short-term maintenance of the SMC differentiation state. They provide evidence that the increased expression of the flow-responsive Klf2 transcription factor in endothelial cells predates SMC differentiation, with the same anterior-to-posterior gradient, and that Klf2 expression is required for SMC differentiation.

Overall, the study is very well-conducted and the paper is well-written. These important data point to hemodynamics as an important driver of artery muscularization in the Circle of Willis.

Reviewer #2 (Public Review):

Summary:

Cheng et al. explore the development of the arteries that form the Circle of Willis and investigate how blood flow pulsatility influences vascular smooth muscle cell (VSMC) differentiation. Using live confocal imaging of the developing zebrafish, the authors show that endothelial cells in the Circle of Willis arteries transition from venous to arterial identity between 54 hours post-fertilization (hpf) and 3 days post-fertilization (dpf), and that this coincides with pdgfrb+ mural cell progenitor differentiation into acta2+ arterial VSMCs. They find that the anterior portions of the Circle of Willis, including the internal carotid arteries (CaDI), establish acta2 expression earlier than posterior aspects, likely due to faster flow rate and increased pulsatility through the CaDI. Then, using computational fluid dynamics, an in vitro co-culture assay, and genetic and drug manipulations of blood flow, the authors provide evidence that pdgfrb+ differentiation is dependent upon pulsatile blood flow and klf2a activation. The results add to our understanding of vascular development and suggest that deficits in pulsatile flow could be potential drivers of arteriopathies.

Strengths:

1. Longitudinal confocal imaging of live developing zebrafish makes the timeline of arterial development in the Circle of Willis easy to understand. This is a strong approach to studying how vascular networks are altered with genetic and pharmacological manipulations.

2. Rigorous use of multiple techniques to test the hypothesis that pulsatile blood flow is required for smooth muscle cell differentiation. The microangiography experiment, in vitro co-culture assay, and genetic and drug manipulations of heart rate at various developmental time points yield outcomes that are consistent with the hypothesis.

Reviewer #3 (Public Review):

Cheng et al. studied if and how blood flow regulates the differentiation of vascular smooth muscle cells (VSMC) in the Circle of Willis (CW) in zebrafish embryos. They show that CW vessels gradually acquire an arterial identity. VSMCs also undergo gradual differentiation, which correlates with blood flow velocity. Using cell culture they show that pulsatile blood flow promotes pericyte differentiation into smooth muscle cells. They further identify transcription factor klf2a as differentially regulated by blood flow, and show that klf2a inhibition results in VSMC differentiation. The authors conclude that pulsatile flow promotes VSMC differentiation through klf2a activation.

Overall this is an important study, because VSMC differentiation in CW has not been previously studied, although analogous observations regarding the role of blood flow and klf2 involvement have been previously made in other systems and other vascular beds, for example, mouse klf2 mutants, which have deficient VSMC coverage of the dorsal aorta (Wu et al., 2008, JBC 283: 3942-50). The results convincingly show that VSMC differentiation in CW depends on the blood flow and that klf2a flow-dependent function regulates VSMC differentiation.

Author response:

The following is the authors’ response to the original reviews.

Weaknesses to be addressed:

(1) More detail is required to understand the effects of genetic and drug manipulations on heart rate as these are important experiments. At the very least, a discussion on the limitations of these manipulations is needed.

- For example, how does one separate the pulsatile versus nutritive effects of blood flow/heartrate reduction?

- The conclusion that arterial SMC differentiation is driven by pulsatile blood flow needs to be toned down. Indeed, this conclusion is mainly supported by in vitro cell co-cultures exposed to laminar versus pulsatile flow. In vivo, reducing Tnnt2a expression affects cardiac contractility and blood flow does not selectively affect pulsatility. To make this conclusion, the authors would need an experimental means to selectively dampen the pulsatility of blood flow.

We understand this concern and we toned down the statements related to the pulsatile flow of our conclusion by using 'flow' instead of 'pulsatile flow' in all text except for the in vitro co-cultures part. We also added a paragraph to discuss the limited capability of qualitatively reduce blood flow in vivo, and acknowledge that the effects of nutrients and flow reduction could not be uncoupled in live zebrafish embryos. We proposed that in the future, in vitro 3D vascular culture models may be combined with microfluidics to precisely calibrate nutrient composition in culture media, flow velocity and pulse; these methods would help address these questions more thoroughly. See page 11-12 line 312-322.

(2) Since mural cells are sensitive to transmural pressure, could the authors elaborate on the potential role of raised intravascular pressure in SMC differentiation? This would better parallel rodents and humans.

We thank you for this suggestion. We added a paragraph to discuss the potential role of raised intravascular pressure in VSMC differentiation in the discussion section (see page 11 line 296-311).

(3) The authors use nifedipine to reduce blood flow. Nifedipine is a specific and potent inhibitor of voltage-dependent calcium channels (VDCC) which are expressed in SMCs. Prior studies (PMID: 35588738) showed that VDCC blockers increased rather than inhibited SMC differentiation. Nifedipine is also likely to act upon VSMC calcium handling in the circle of Willis, which may in turn affect cell maturation. Could the authors comment on this seeming discrepancy?

It is possible that off-target or indirect effects of Nifedipine decrease smooth muscle cell proliferation, or that altered cardiac contractility fundamentally alters aspects of vascular development other than blood flow.

- Additionally, it would be helpful to report the quantitative heart rate reduction achieved with Nifedipine. This would clear up concerns that the heart rate reduction is too large for normal vascular development to occur, and thus decrease proliferation rate independent of changes in blood flow pulsatility.

We concur with these comments, which is why our experimentation with Nifedipine is reinforced by employing an alternative, non-pharmacological strategy to inhibit blood flow: the use of morpholino against tnnt2a gene. The results with either Nifedipine or tnnt2a support the lack of VSMCs maturation. In addition, we provided the quantitative heart rate reduction achieved with Nifedipine shown in new Figure S2A-S2C, suggesting that the drug is not completely halting the heart rate but decreasing it. Nevertheless, we report that Zebrafish embryos can survive and develop a normal blood vascular system without any heartbeat. Hence, we exclude that the effect on VSMCs maturation is linked non-specifical effects caused by the loss of heartbeat. Nevertheless, we now acknowledged in our discussion the limitation of nifedipine, as it may affect VSMC through VDCCs (page 12, line 323-334).

We also added a paragraph in the discussion section to compare nifedipine, an L-type VDCC blocker, and ML218, a T-type VDCC selective inhibitor from the previous study (Ando et al., 2022). We noted that in this previous study, the increase in VSMC differentiation only occur on anterior metencephalic central arteries (AMCtAs) that are more than 40 mm away from the BCA; these AMCtAs are much smaller than CoW arteries and have different geometry hence possible different kinetics of VSMC maturation (Ando et al., 2022) as our manuscript discovery would suggest.

(4) The authors should provide more information on how blood flow velocity and wall shear stress are calculated from the Circle of Willis vascular structure. It is presumed that these values are dependent upon the 3-D morphology of the vessel network, as labeled by intravenous dextran dye, but this is not clear. (a second reviewer similarly comments: I was unclear how flow velocity values were obtained in Fig. 3E. Are they based on computational simulation, or are they experimentally calculated following the dextran injection?) Small local differences in vessel diameter and shape will influence blood flow velocity, but these morphological changes are not clearly articulated. Further, it is unclear how flow input levels to the CaDI and basilar arteries are decided across time points. For instance, is it possible to measure the blood flow speed empirically with line-scanning or high-speed tracking of labeled blood cells or particles? This would provide validation of the modeling results.

The computational fluid dynamic simulation was performed according to previous study from our lab (Barak et al., 2021). Blood flow velocity and wall shear stress are dependent upon the 3D morphology of the vessel network labeled by intravascular dextran. Details on how the computational fluid dynamic simulation was performed are added in method section page 17 line 433-449.

Moreover, to address this reviewer concern we have now provided new experimental measurement of blood flow using the red blood cell (RBC) velocity via axial line scanning microscopy in Tg(kdrl:gfp;gata1:DsRed)zn1/sd2 zebrafish embryos at 54 hpf, 3 dpf, and 4 dpf. By using the experimental RBC velocity, we re-simulated the computational fluid dynamic. The new findings align with our conclusion and are further elaborated upon in response to this reviewer comment listed as point 6. Details on how RBC velocity calculated is added in method section page 16 line 414-431.

(5) Does the cardiac injection of dextran itself affect the diameter of the arteries, given the invasiveness of the procedure? This could be examined in fish with a transgenic endothelial label with and without dextran.

Here, we performed an experiment on wildtype zebrafish at 5 days post-fertilization (dpf) with and without Dextran injection, examining the effects of Dextran injection on vessel diameters. As shown in the representative image below, the XZ panel clearly illustrates a Dextran-filled PCS vessel with no alteration in vessel size. Dextran microangiography, a technique employed to obtain vessel geometry with fluorescent microsphere, has been well established in zebrafish (Kamei et al., 2010). Our findings, demonstrating that Dextran does not affect vessel size, are consistent with previous studies utilizing Dextran microangiography.

(6) The data from the microangiography experiment in Figure 3 does not fully support the stated results. The authors report that the CaDI had the highest blood flow speed starting from 54 hpf, but it does not appear to be higher than the other arteries at this time point. Additionally, there is not sufficient evidence that wall shear stress coincides with smooth muscle cell differentiation in the CaDI. Wall shear stress appears to be similar between 54 hpf and 3 dpf in the CaDI, only increasing between 3 dpf and 4 dpf, while differentiation is shown to begin at 3 dpf. The authors need to address this and/or soften conclusions.

First, In response to this specific reviewer concern, we measured red blood cell (RBC) velocity by used axial line scanning microscopy to analyze Tg(kdrl:gfp;gata1:DsRed)zn1/sd2 zebrafish embryos (the detailed method was added in Method section in the manuscript). We replaced the computational simulated blood flow velocity by RBC velocity in new Figure 3E-3G, and re-run the computational simulated wall shear stress (WSS) using the RBC velocity in new Figure 3I-3K. We compared RBC velocity and WSS among different vessels at each time point. We confirmed that CaDI has the highest RBC velocity starting from 54 hpf to 4 dpf (new Figure 3A-3C, and 3E-3G) and found an overall increase in average WSS from 54 hpf to 4 dpf (new Figure 3A-3C, and 3H). Further, WSS in CaDI was significantly higher than BCA and PCS at 54 hpf, 3 dpf, and 4 dpf (new Figure 3A-3C, 3I-3K). Altogether, the CFD simulation suggests that CoW arteries experience different hemodynamic WSS that is associated with spatiotemporal pattern of VSMC differentiation on CoW arteries.”. (Page 6, line 153-162)

Second, to identify the correlation of WSS and VSMC differentiation in CaDI, we performed Pearson correlation analysis. In the image provided here, we plotted a linear regression with normalized # of acta2+ cells in CaDI and WSS with developmental stages (54 hpf, 3 and 4 dpf), and performed Pearson correlation coefficient analysis by using GraphPad Prism 10.0.3. The correlation coefficient r = 0.595, suggesting that the two variables (acta2+ cells and WSS) tend to increase together with developmental stages (54 hpf, 3 and 4 dpf).

Third, we softened our conclusion as the RBC velocity across CoW arteries was differentially distributed while VSMC differentiation occurred in these vessels.

(7) It is unclear if acta2 expression is conferring vascular tone, as would be expected if the cells are behaving as mature VSMCs. Does arterial diameter decrease with an increase in acta2 expression? Are acta2-positive mural cells associated with more dynamic changes in arteriole diameter under basal or stimulated conditions?

Thanks for this interesting question. VSMC maturation and its vasoactivity could be further investigated in the future. Our study focused on early stage of VSMC differentiation, in which pdgfrb+ progenitors started to express VSMC marker acta2. We discussed the onset of transgelin expression and loss of abcc9 expression as markers of VSMC maturation. In addition, a previous study found that VSMC covered vessels in zebrafish brain dilate as early as 4 dpf and constrict at 6 dpf (Bahrami & Childs, 2020). Future study may focus on the association between expression of different VSMC markers and VSMC functional maturation. (page 10, line 272-279)

(8) The authors argue that CoW vessels transition from venous to arterial identity (Fig. 1). However, kdrl is not an ideal arterial marker for this experiment as it is expressed in both arteries and veins. While it is true that many arterial beds have stronger kdrl expression than the veins, its expression in both arteries and veins changes with developmental stage, and its expression level may vary depending on the type of vessel. Therefore, showing that kdrl increases from 32 hpf - 4 dpf in CoW vessels is not convincing because its expression may increase in both venous or arterial vasculature as the vessels mature. In addition, flt4 expression is not exclusively venous; for example, it has noticeable expression in the dorsal aorta at 24-32 hpf stages. It would be helpful to confirm this transition by analyzing additional arterial and venous markers.

We acknowledge this and we added a paragraph to discuss the limitation. We combined loss of flt4 and increase in kdrl to establish the temporal sequence of circle of Willis morphogenesis, arterial specification, and VSMC differentiation. We acknowledge that additional arterial and venous markers need to be analyzed for a more thorough characterization of arterial specification in vertebrate brain vascular development. See page 12 line 335-341.

(9) The authors show that acta2+ VSMCs are absent in tnnt2a MO embryos, concluding that blood flow is required for their differentiation from pericytes. However, there is no data showing that pericytes are still present in tnnt2a MO embryos. Although this has been previously shown by Ando et al 2016, it would be beneficial to confirm in the current study as this is a critical piece of evidence needed for this conclusion.

To determine if blood flow is dispensable for pdgfrb+ progenitor recruitment, we performed tnnt2a MO (0.35 ng/embryo) injection in Tg(pdgrb:egfp, kdrl:ras-mcherry) ncv22/s896. Loss of blood flow did not affect pdgfrb+ progenitor emergence around the CoW (new Figure S2G-S2H) at 3 days post fertilization (dpf). This is consistent with previous observation in Ando et al 2016 Figure S2C (Ando et al., 2016).

(10) The authors show that klf2a MO injected embryos have a reduced number of VSMCs at 3 dpf but a normal number at 4 dpf (Fig. 6), concluding that klf2a is only important to initiate CaDI muscularization. If this is true, it would raise important questions about how VSMCs differentiate at a later stage in the absence of klf2a. For instance, is blood flow not required to differentiate at a later stage, or is there another factor that compensates in the absence of klf2a? The alternative explanation/ caveat is that klf2a MO loses efficacy with development, leading to the recovery of VSMCs at this stage. Therefore, it would be important to confirm this result using a genetic klf2a mutant.

Thank you for pointing this out. We note that based on the klf2a reporter line, klf2a activity in CoW arterial endothelial cells is highly correlated with the number of acta2+ VSMCs in CaDI, BCA and PCS at 3 dpf (r = 0.974, new Figure S5J). Interestingly however, klf2a activity remained stable from 3 dpf to 4 dpf, well beyond initiation of VSMC differentiation. Thus, we speculate sustained klf2a expression may support further maturation of VSMCs, as acta2+ VSMCs showed distinct morphology at 4 dpf compared with 3 dpf. (Page 10, line 268-272). As for the observation that klf2a morphants have normal number of VSMCs at 4 dpf, we think that in addition to the temporary effect of morpholino, a proximal explanation is compensation by paralogous klf2b in zebrafish. We acknowledge that further characterization of CoW VSMC development in klf2a and klf2b double genetic mutants (Rasouli et al., 2018; Steed et al., 2016) may help determine whether klf2b compensates klf2a in CoW VSMC differentiation beyond 4 dpf. See page 10-11 line 292-295.

(11) A large part of the discussion focuses on Notch and Wnt signaling, as downstream Klf2 effectors. While these are reasonable hypotheses to propose, there is no data on the involvement of these pathways in the current study. It seems excessive to speculate on detailed mechanisms of how Klf2 activates Notch and Wnt signaling in the absence of data showing that these pathways are affected in CoW vessels. Therefore, the discussion could be shortened here unless additional data can be obtained to demonstrate the involvement of these pathways in VSMCs in CoW.

We concur and have condensed the discussion on Notch and Wnt signaling as downstream klf2 effectors.

Minor comments:

(1) Line 138 "CaDI is the only vessels in the CoW receiving pulsatile arterial blood low ... ". Adding a reference to support this statement would be useful.

We agree and revised this sentence into ‘CaDI receive proximal arterial feed through lateral dorsal aorta from cardiac outflow tract (Isogai et al., 2001)’. It was also based on our general observation of zebrafish vascular anatomy and blood flow under a confocal microscope.

(2) The image insets in Figs. 1A, 2A, 4E-L, 5A, 6A are quite small. Please make them larger to help the reader interpret the findings.

We agree. We maximized the image size to help the reader interpret the finding, and to visualize confocal images and schematics side-by-side.

(3) The schematics in Figs. 1-2, and 4-6 are helpful, but the different cell types are difficult to see because they are small and their colors/shapes are not very distinct.

We agree. We increased the size and color contrast to provide better visualization of the schematics in new schematic Figures. 1-2 and 4-6.

(4) It is stated that there are no diameter differences between different arteries, but statistics are not reported.

The statistics in Figure 3D were performed by ordinary two-way ANOVA followed by Tukey’s multiple comparisons test, with a single pooled variance. Here we added pairwise comparisons among vessels in the CoW. Hence when non indicated the difference are non-significant.

(5) Figure 3F would be better visualized on a log scale, as it is difficult to see the differences between each post-fertilization timepoint.

We agree. In the new Figure 3H, the average wall shear stress (WSS) in CoW arteries is presented on log scale in y axis to see the differences between each post-fertilization timepoint.

(6) Please provide more background and validation on the pericyte cell line, and their use for the questions in this study.

Thank you for the question, TgBAC(pdgfrb:egfp)ncv22 was generated and described by Ando et al 2016 to clarify mural cell coverage of vascular endothelium in zebrafish (Ando et al., 2016). We added a describe in the method section to provide background and validation on this pericyte line (see page 13 line 368-372).

(7) Flow velocity and WSS changes are shown in each vessel in Figs. 3E,G. However, the comparison should be made between different types of vessels to see if there is a statistical difference and PCS, for example, which would explain differences in VSMC coverage.

We agreed. We compared the difference among arteries in the CoW at each developmental timepoint and performed ordinary one-way ANOVA with Tukey’s multiple comparisons test. Figure. 3E is replaced by new Figure. 3E-G and Figure. 3G is replaced by new Figure. 3I-K.

(8) Similarly, between CaDI, the number of klf2a cells in Fig. 5B should be compared between different vessels, not between different stages of the same vessel.

We agree. In new Figure 5B-E, the number of klf2a+ cells per 100 μm vessel length are compared among different vessels at each developmental stage and analyzed by ordinary one-way ANOVA with Tukey’s multiple comparisons test.

(9) When quantifying klf2+ cells in Fig. 5, it would be helpful to quantify klf2 expression level between cells in different vessels. This could be done by quantifying GFP expression in existing images. The difference in expression level may explain the variation between CaDI and PCS more accurately than just the difference in cell number.

The GFP expression reflect the stability of GFP protein expression and labels discrete nuclei with active klf2a expression. Hence the quantification of GFP level might not give an accurate readout of klf2a expression per se but rather of its activity. For this reason we don’t think that this experiment will add accurate measurement of klf2a expression.

(10) Do data points in Figure 4D correspond to different cells in the same chamber experiment? If so, they cannot be treated as independent replicates. Each data point should correspond to an independent replicate experiment.

We agree. Now in the figure legend, we report the number of cells analyzed.

(11) Graph placement is confusing in Figs. 4I, M. An adjacent Fig. 4G shows Nifedipine treated embryos, while the graph next to (Fig. 4I) shows acta+ cell number from tnnt2a 4 dpf experiment. Similarly, the bottom Fig. 4K tnn2a 4 dpf MO experiment has an adjacent graph Fig. 4M, which shows nifedipine treatment quantification, which makes it very confusing.

We agreed. We rearranged Figure 4E (representative images of control embryos at 3 dpf and 4 dpf), Figure 4F (tnnt2a MO embryos at 3 dpf and 4 dpf), Figure 4G (nifedipine treated embryos at 3 dpf and 4 dpf).

Reference:

Ando, K., Fukuhara, S., Izumi, N., Nakajima, H., Fukui, H., Kelsh, R. N., & Mochizuki, N. (2016). Clarification of mural cell coverage of vascular endothelial cells by live imaging of zebrafish. Development, 143(8), 1328-1339. https://doi.org/10.1242/dev.132654

Ando, K., Tong, L., Peng, D., Vazquez-Liebanas, E., Chiyoda, H., He, L., Liu, J., Kawakami, K., Mochizuki, N., Fukuhara, S., Grutzendler, J., & Betsholtz, C. (2022). KCNJ8/ABCC9-containing K-ATP channel modulates brain vascular smooth muscle development and neurovascular coupling. Dev Cell, 57(11), 1383-1399 e1387. https://doi.org/10.1016/j.devcel.2022.04.019

Bahrami, N., & Childs, S. J. (2020). Development of vascular regulation in the zebrafish embryo. Development, 147(10). https://doi.org/10.1242/dev.183061

Barak, T., Ristori, E., Ercan-Sencicek, A. G., Miyagishima, D. F., Nelson-Williams, C., Dong, W., Jin, S. C., Prendergast, A., Armero, W., Henegariu, O., Erson-Omay, E. Z., Harmanci, A. S., Guy, M., Gultekin, B., Kilic, D., Rai, D. K., Goc, N., Aguilera, S. M., Gulez, B., . . . Gunel, M. (2021). PPIL4 is essential for brain angiogenesis and implicated in intracranial aneurysms in humans. Nat Med, 27(12), 2165-2175. https://doi.org/10.1038/s41591-021-01572-7

Isogai, S., Horiguchi, M., & Weinstein, B. M. (2001). The vascular anatomy of the developing zebrafish: an atlas of embryonic and early larval development. Dev Biol, 230(2), 278-301. https://doi.org/10.1006/dbio.2000.9995

Kamei, M., Isogai, S., Pan, W., & Weinstein, B. M. (2010). Imaging blood vessels in the zebrafish. In Methods in cell biology (Vol. 100, pp. 27-54). Elsevier.

Rasouli, S. J., El-Brolosy, M., Tsedeke, A. T., Bensimon-Brito, A., Ghanbari, P., Maischein, H. M., Kuenne, C., & Stainier, D. Y. (2018). The flow responsive transcription factor Klf2 is required for myocardial wall integrity by modulating Fgf signaling. Elife, 7. https://doi.org/10.7554/eLife.38889

Steed, E., Faggianelli, N., Roth, S., Ramspacher, C., Concordet, J. P., & Vermot, J. (2016). klf2a couples mechanotransduction and zebrafish valve morphogenesis through fibronectin synthesis. Nat Commun, 7, 11646. https://doi.org/10.1038/ncomms11646

Version 2

Abstract

Abstract

Vascular smooth muscle cells (VSMCs) envelop vertebrate brain arteries, playing a crucial role in regulating cerebral blood flow and neurovascular coupling. The dedifferentiation of VSMCs is implicated in cerebrovascular diseases and neurodegeneration. Despite its importance, the process of VSMC differentiation on brain arteries during development remains inadequately characterized. Understanding this process could aid in reprogramming and regenerating differentiated VSMCs in cerebrovascular diseases. In this study, we investigated VSMC differentiation on the zebrafish circle of Willis (CoW), comprising major arteries that supply blood to the vertebrate brain. We observed that the arterial expression of CoW endothelial cells (ECs) occurs after their migration from the cranial venous plexus to form CoW arteries. Subsequently, acta2+ VSMCs differentiate from pdgfrb+ mural cell progenitors upon recruitment to CoW arteries. The progression of VSMC differentiation exhibits a spatiotemporal pattern, advancing from anterior to posterior CoW arteries. Analysis of blood flow suggests that earlier VSMC differentiation in anterior CoW arteries correlates with higher red blood cell velocity wall shear stress. Furthermore, pulsatile blood flow is required for differentiation of human brain pdgfrb+ mural cells into VSMCs as well as VSMC differentiation on zebrafish CoW arteries. Consistently, the flow-responsive transcription factor klf2a is activated in ECs of CoW arteries prior to VSMC differentiation, and klf2a knockdown delays VSMC differentiation on anterior CoW arteries. In summary, our findings highlight the role of blood flow activation of endothelial klf2a as a mechanism regulating the initial VSMC differentiation on vertebrate brain arteries.

Introduction

Vascular smooth muscle cells (VSMCs) are contractile mural cells wrapping around endothelial cells (ECs) of large vessels, especially arteries (Ando et al., 2022; Donadon & Santoro, 2021; Stratman et al., 2020). The expression of contractile proteins, such as alpha-smooth muscle actin (encoded by acta2), distinguishes VSMCs from pericytes, the other type of mural cells primarily associated with small vessels (Bahrami & Childs, 2020; Donadon & Santoro, 2021; Stratman et al., 2020). VSMCs are important for arterial homeostasis (Basatemur, Jorgensen, Clarke, Bennett, & Mallat, 2019). In the brain, VSMC constriction and relaxation is essential for functional hyperemia in neurovascular coupling (Hill et al., 2015; Kaplan, Chow, & Gu, 2020). Phenotype switching or dedifferentiation of VSMCs, marked by lower expression of contractile proteins, is prominent in pathological progression of atherosclerosis (Basatemur et al., 2019). VSMC phenotype switching or dedifferentiation also contributes to cerebrovascular diseases and neurodegeneration (Aguilar-Pineda et al., 2021; Chou et al., 2022; Frosen & Joutel, 2018; Milewicz et al., 2010; Oka et al., 2020; Poittevin et al., 2014). As development and regeneration share many common molecular and cellular mechanisms (Nowotarski & Sánchez Alvarado, 2016), insights on VSMC differentiation on brain arteries during development may inform strategies to reprogram and regenerate dedifferentiated VSMCs and alleviate cerebrovascular diseases.

VSMCs differentiate from heterogenous perivascular progenitors of mesoderm and neural crest origins (Ando et al., 2019; Donadon & Santoro, 2021; Whitesell et al., 2019). Previous research described various mechanisms that regulate VSMC differentiation. Autonomous Notch activation is required for specification of pdgfrb+ mural cell progenitors from peri-arterial mesenchyme, and these progenitors later differentiate into acta2+ VSMCs (Ando et al., 2019). Arterial Notch signaling activated by blood flow is necessary for acta2+ VSMC emergence on zebrafish trunk dorsal aorta (X. Chen, Gays, Milia, & Santoro, 2017). Chemokine signaling promotes VSMC association with zebrafish dorsal aorta, whereas blood flow responsive transcription factor krüppel-like factor 2 (encoded by klf2a in zebrafish) prevents their association with the adjacent cardinal vein (Stratman et al., 2020). Endothelial BMP signaling and autonomous foxc1 expression regulate VSMC differentiation in zebrafish ventral head (Watterston, Zeng, Onabadejo, & Childs, 2019; Whitesell et al., 2019). Autonomous expression of an ATP-sensitive potassium channel modulates VSMC differentiation from pdgfrb+ progenitors in the brain (Ando et al., 2022). These studies suggest that VSMC differentiation is highly organotypic.

The circle of Willis (CoW) consists of major arteries that supply blood to the vertebrate brain, including internal carotid arteries (ICAs) and posterior communicating arteries (PCAs) (Campbell et al., 2019; Schröder, Moser, & Huggenberger, 2020). In early development, ICAs are the first large arteries to form and supply all circulation in the developing brain (Kathuria, Gregg, Chen, & Gandhi, 2011; Menshawi, Mohr, & Gutierrez, 2015). In adults, posterior circulation becomes independent, and ICAs remain major feeding arteries in anterior, while connected to posterior circulation with PCAs, completing a circle of arteries (Kathuria et al., 2011; Schröder et al., 2020). In adults, intracranial aneurysms are frequently found close to CoW arteries (Frosen & Joutel, 2018). CoW arteries are wrapped by VSMCs, and VSMC dedifferentiation and hyperplasia are described in carotid atherosclerosis and Moyamoya disease (Chou et al., 2022; Fox, Dorschel, Lawton, & Wanebo, 2021). Spatiotemporal dynamics of VSMC differentiation on vertebrate CoW arteries during development have not been thoroughly investigated. VSMC differentiation is difficult to observe in vivo with mammalian models, as mammalian embryos develop in utero and depend on maternal circulation. Zebrafish is an important vertebrate model for vascular development, as zebrafish embryos are externally fertilized, optically clear, and thus accessible to confocal live imaging of developing blood vessels (Stratman & Weinstein, 2021). Establishment of zebrafish fluorescent transgenic reporter lines with pdgfrb and acta2 promoters enables in vivo visualization and assessment of mural cells and VSMCs (Ando et al., 2016; Whitesell et al., 2014). In addition, zebrafish CoW shows comparable geometry to human CoW: the caudal division of the internal carotid artery (CaDI) in zebrafish is comparable to human ICA, and the posterior communicating segment (PCS) in zebrafish is comparable to human PCA (Isogai, Horiguchi, & Weinstein, 2001). Taking advantage of the zebrafish model, we describe the spatiotemporal pattern of VSMC differentiation on CoW arteries, and associate this pattern with blood flow activation of klf2a in arterial ECs.

Results

Artery muscularization is spatiotemporally regulated in CoW arteries

In zebrafish brain, arteries are formed by ECs from cranial venous plexus (Fujita et al., 2011). Recent single-cell transcriptome profiling of vascular cells in prenatal human brain also supports an endothelial differentiation trajectory from proliferative venous ECs to quiescent arterial ECs (Crouch et al., 2022). To establish the temporal sequence of CoW arterial specification and VSMC differentiation, we started with confocal live imaging of fluorescent transgenic reporter lines Tg(flt4:yfp)hu4881, which labels primitive venous ECs, and Tg(kdrl:hras-mcherry)s896, which is expressed in all ECs but enriched in arterial ECs, at four stages starting at 32 hours post fertilization (hpf), when CoW arteries are assembled (Chi et al., 2008; Hogan et al., 2009; Isogai et al., 2001). As arterial ECs express higher kdrl, we used the intensity of mcherry as an indicator of artery specification (Figure 1A-1E) (Chi et al., 2008). We found that when ECs from cranial venous plexus migrated to form CoW arteries at 32 hpf, they still retained primitive venous identity, as indicated by flt4 expression (Figure 1A). When connections between cranial venous plexus and CoW arteries disappeared at 54 hpf, CoW arteries no longer expressed primitive venous flt4 (Figure 1B). Expression of arterial enriched kdrl in CaDI and PCS at 3 days post fertilization (dpf) was significantly higher than 32 hpf (Figure 1C and 1E). At 4 dpf, kdrl expression in CoW arteries remained comparable to 3 dpf, suggesting that arterial specification of CoW was completed by 3 dpf (Figure 1C-1E). Together, these data suggest CoW morphogenesis precedes arterial endothelial gene expression.

Artery gene expression of endothelial cells (ECs) in circle of Willis (CoW) arteries

(A-D) Confocal live images of Tg(flt4:yfp, kdrl:hras-mcherry)hu4881/s896 and scheme representation of CoW arteries in zebrafish brain at 32 hour post fertilization (hpf) (A), 54 hpf (B), 3 day post fertilization (dpf) (C), and 4 dpf (D). Green channel represents flt4:yfp fluorescence, Red channel represents kdrl:hras-mcherry fluorescence, and Merge panel combines both channels. Arrows point to the CoW arteries with kdrl:hras-mcherry signal. Scale bar = 50 μm

(E) Average intensity of kdrl:hras-mcherry in caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=6, 2 independent experiments), 54 hpf (n=12, 4 independent experiments), 3 dpf (n=15, 4 independent experiments), and 4 dpf (n=9, 2 independent experiments), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, * p≤0.05, ** p≤0.01

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Previous research suggested that acta2+ VSMCs on CoW arteries differentiate from pdgfrb+ mural cell progenitors (Ando et al., 2022; Ando et al., 2019). Thus, to characterize spatiotemporal dynamics of VSMC differentiation on CoW arteries, we performed confocal live imaging of Tg(acta2:mcherry)ca8, TgBAC(pdgfrb:egfp)ncv22, and Tg(kdrl:cerulean)sd24, to visualize respectively VSMCs (red), pdgfrb+ progenitors (green), and developing arteries (white) at four stages starting from 32 hpf (Figure 2) (Ando et al., 2016; Page et al., 2013; Whitesell et al., 2014). We found that pdgfrb+ progenitors emerged on CoW arteries around 54 hpf after CoW morphogenesis, and most pdgfrb+ progenitors did not express acta2 at 54 hpf (Figure 2A-2B, 2E-2F, and S1A-S1B). The number of pdgfrb+ progenitors on CoW arteries increased temporally from 54 hpf to 3 dpf (Figure 2B, 2C, and 2E). Notably, while many pdgfrb+ cells on CaDI differentiated into acta2+ VSMCs by 3 dpf, acta2+ VSMCs were hardly observed on adjacent BCA or PCS, suggesting that VSMC differentiation started between 54 hpf and 3 dpf on CaDI but not BCA or PCS (Figure 2B-2C, 2E-2F, and S1A-S1B). At 4 dpf, the number of pdgfrb+ cells on CoW arteries remained comparable to 3 dpf, while the number of acta2+ VSMCs on CaDI increased (Figure 2C-2F and S1A-S1B). Importantly, most pdgfrb+ cells on BCA and PCS only started to express acta2 around 4 dpf, much later than CaDI (Figure 2C-2F and S1A-S1B). These findings support that differentiation of acta2+ VSMCs from pdgfrb+ progenitors progresses spatially and temporally along the anterior to posterior direction in CoW arteries.

Vascular smooth muscle cell (VSMC) differentiation on circle of Willis (CoW) arteries

(A-D) Confocal live images of Tg(acta2:mcherry, kdrl:cerulean)ca8/sd24 TgBAC(pdgfrb:egfp)ncv22 and scheme representation of vascular endothelium and mural cells on CoW arteries in zebrafish brain at 32 hour post fertilization (hpf) (A), 54 hpf (B), 3 day post fertilization (dpf) (C), and 4 dpf (D). White channel represents kdrl:cerulean fluorescence, Red channel represents acta2:mcherry, Green channel represents pdgfrb:egfp, Merge 1 panel combines all three channels, Merge 2 combines acta2:mcherry in red and pdgfrb:egfp in green. Arrows point to the CoW arteries with pdgfrb:egfp and acta2:mcherry signal. Scale bar = 50 μm

(E) Number of pdgfrb+ vascular mural cell progenitors per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=3, 1 independent experiment), 54 hpf (n=6, 1 independent experiment), 3 dpf (n=6, 1 independent experiment), and 4 dpf (n=5, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, ** p≤0.01, *** p≤0.001, **** p≤0.0001

(F) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 32 hpf (n=5, 1 independent experiment), 54 hpf (n=6, 2 independent experiments), 3 dpf (n=24, 6 independent experiments), and 4 dpf (n=25, 6 independent experiments), two-way analyses of variance followed by Tukey’s multiple comparisons, represented with mean ± SD, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

CoW arteries have spatiotemporal difference in hemodynamics

Our data suggest that VSMCs on CaDI of anterior CoW differentiate much earlier than BCA and PCS of posterior CoW, although all acta2+ VSMCs on CoW differentiated from pdgfrb+ progenitors after they were recruited to CoW (Figure 2 and S1). In addition, anterior and posterior CoW vascular morphogenesis and arterial specification occurred around the same time (Figure 1). We speculate that hemodynamic distribution may contribute to the spatiotemporal difference in VSMC differentiation on CoW arteries. CaDI receive proximal arterial feed through lateral dorsal aorta from cardiac outflow tract (Isogai et al., 2001). Furthermore, CaDI differs from the BCA or PCS in terms of overall vascular geometry, which is among determinants of local vascular hemodynamic forces such as wall shear stress (WSS) (Katritsis et al., 2007). To explore this possibility, we performed microangiography with a high molecular weight fluorescent dextran to determine each CoW artery diameter. We found that there was no significant different among CaDI, BCA and PCS (Figure 3A-3D). Then, we measure red blood cell (RBC) velocity by axial line scanning microscopy to analyze Tg(kdrl:gfp;gata1:DsRed)zn1/sd2 zebrafish at 54 hpf, 3 and 4 dpf. The RBC velocity across CoW arteries was differentially distributed, with CaDI having the highest RBC velocity from 54 hpf to 4 dpf (Figure 3E-3G). This is consistent with CaDI being the inlet vessel for arterial blood flow within the entire CoW circulation. Next, we used computational fluid dynamics (CFD) to simulate the effect of flow on CoW, based on the Dextran-filled CoW geometric network and RBC velocity (ANSYS, 2014; Z. Chen et al., 2019; Fernandes et al., 2022). We found an overall increase in average WSS from 54 hpf to 4 dpf (Figure 3A-3C, and 3H). Further, the simulated flow pattern and WSS on the CoW showed that WSS in CaDI was significantly higher than BCA and PCS at 54 hpf, 3 dpf, and 4 dpf (Figure 3A-3C, 3I-3K). Moreover, upon comparing WSS with the number of acta2+ VSMCs per CoW arteries during the development, we observed a Pearson correlation coefficient of r =0.595, indicating a moderate to strong positive correlation between these two variables (Figure S1C). In summary, the analysis of blood flow suggests that CoW arteries undergo varied hemodynamic WSS, which correlates with the spatial and temporal pattern of VSMC differentiation along CoW arteries.

Computational fluid dynamic (CFD) simulation of circle of Willis (CoW) arteries

(A-C) Confocal live images of Tg(kdrl:gfp)zn1 injected with dextran and representation of vascular diameter, CFD simulated flow and wall shear stress (WSS) within the CoW arteries at 54 hour post fertilization (hpf) (A), 3 day post fertilization (dpf) (B), and 4 dpf

(C). Arrows point to the CoW arteries with high flow velocity and WSS. Scale bar = 50 μm

(D) Average vascular diameter in caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 54 hpf (n=6, 1 independent experiment), 3 dpf (n=6, 1 independent experiment), and 4 dpf (n=6, 1 independent experiment), two-way analyses of variance followed by Tukey’s multiple comparisons, *** p≤0.001

(E-G) Red blood cell (RBC) velocity in CaDI, BCA, and PCS at 54 hpf (n=4, 1 independent experiment) (E), 3 dpf (n=5, 1 independent experiment) (F), and 4 dpf (n=4, 1 independent experiment) (G), ordinary one-way ANOVA analyses with Tukey’s multiple comparisons, represented with mean ± SD, * p≤0.05, ** p≤0.01

(H) Average wall shear stress (WSS) throughout the CoW arteries at 54 hpf (n=3, 1 independent experiment), 3 dpf (n=3, 1 independent experiment), and 4 dpf (n=3, 1 independent experiment), ordinary one-way ANOVA analyses with Tukey’s multiple comparisons, **** p≤0.0001

(I-K) Average wall shear stress (WSS) in CaDI, BCA, and PCS at 54 hpf (n=3, 1 independent experiment) (I), 3 dpf (n=3, 1 independent experiment) (J), and 4 dpf (n=3, 1 independent experiment) (K), ordinary one-way ANOVA analyses with Tukey’s multiple comparisons, represented with mean ± SD, * p≤0.05, ** p≤0.01, *** p≤0.001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, RBC: red blood cell, WSS: wall shear stress, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Blood flow is required for CoW artery muscularization

Our data suggest that ECs under arterial flow might favor acta2+ VSMC differentiation from pdgfrb+ progenitors. To test this hypothesis, we set up an in vitro cell co-culture experiment where GFP-PDGFRB+ human brain pericytes (GFP-HBVPs) were cultured in flow amenable outlet slides, then covered by a thin layer of collagen type 1. Confluent human endothelial cells (HUVECs) were seeded on top of that and then exposed to steady-state laminar or pulsatile flow conditions (Figure 4A) (Abello, Raghavan, Yien, & Stratman, 2022). Differentiation of PDGFRB+ human brain pericytes into ACTA2+ VSMCs was then quantified after 24 hours of flow treatment. We found that brain pericytes under pulsatile flow were larger and showed higher expression of VSMC differentiation markers including PDGFRB, ACTA2, and TRANSGELIN (Robin et al., 2013) (Figure 4B-4D), suggesting that pulsatile flow favored PDGFRB+ progenitor differentiation into VSMC.

Blood flow is required for vascular smooth muscle cell (VSMC) differentiation on circle of Willis (CoW) arteries

(A) Scheme representation of in vitro cell co-culture experiment

(B) Representative immunofluorescence images of brain pericytes after exposure of pulsative flow and laminar flow. Cells were stained for ACTA2 (cyan), and cytosolic GFP label (magenta). Scale bar = 10 μm

(C) Morphological measurement of brain pericyte length/width ratio before and after exposure of pulsative flow and laminar flow (n = 63 cells), two-tailed Mann–Whitney test, represented with mean ± SD, **** p≤0.0001

(D) Protein level of PDGFRB (n = 153 cells), ACTA2 (n = 222 cells), and TAGLN (n = 150 cells) in arbitrary unit (A.U.) after exposure of pulsative flow and laminar flow, two-tailed Mann–Whitney test, represented with mean ± SD, **** p≤0.0001

(E-G) Confocal live images of Tg(acta2:mcherry; kdrl:gfp)ca8/zn1 and scheme representation of vascular endothelium and VSMCs on CoW arteries at 3 day post fertilization (dpf) and 4 dpf in control embryos (E), embryos injected with 0.35 ng tnnt2a morpholino (MO) (F), and embryos treated with 25 μM nifedipine from 54 hpf (G). Red channel represents acta2:mcherry, Green channel represents kdrl:gfp, and Merge panel combines both channels. Arrows point to the CoW arteries with acta2:mcherry signal. Scale bar = 50 μm

(H) Number of acta2+ VSMCs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 3 dpf in uninjected control (n=6, 1 independent experiment) and embryos injected with 0.35 ng tnnt2a MO at one to two cell stage (n=6, 1 independent experiment), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, *** p≤0.001, **** p≤0.0001

(I) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in uninjected control (n=6, 1 independent experiment) and embryos injected with 0.35 ng tnnt2a MO at one to two cell stage (n=6, 1 independent experiment), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, **** p≤0.0001

(I) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 3 dpf in DMSO control (n=8, 2 independent experiments) and embryos treated with 25 μM nifedipine from 54 hpf (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01, **** p≤0.0001

(J) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in DMSO control (n=9, 2 independent experiments) and embryos treated with 25 μM nifedipine from 54 hpf (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, MO: morpholino, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Previous research suggests that blood flow is dispensable for recruitment of pdgfrb+ progenitors on BCA and PCS (Ando et al., 2016). To corroborate this, we injected 0.35 ng of tnnt2a morpholino antisense oligonucleotide (MO) into Tg(pdgfrb:egfp, kdrl:hras-mcherry)ncv22/s896 at one to two-cell stage, to abrogate cardiac contractility and thus blood flow (Ando et al., 2016; Chi et al., 2008; Sehnert et al., 2002). Indeed, the number of pdgfrb+ cells on CaDI, BCA, and PCS at 3 dpf in tnnt2a morphants was comparable to uninjected sibling control, suggesting blood flow did not affect pdgfrb+ progenitor recruitment to CoW arteries (Figure S2G-S2H). To determine the effect of blood flow on VSMC differentiation, we injected tnnt2a MO into Tg(acta2:mcherry, kdrl:gfp)ca8/zn1 (Cross, Cook, Lin, Chen, & Rubinstein, 2003; Sehnert et al., 2002; Whitesell et al., 2014). Notably, we found no acta2+ VSMCs on CoW arteries of tnnt2a morphants at either 3 or 4 dpf (Figure 4E-4F and 4H-4I), suggesting that blood flow was required for acta2+ VSMC differentiation on CoW arteries.

To further explore the temporal requirement of blood flow, we treated embryos with 25 μM nifedipine, a drug shown to reduce heart rate in zebrafish embryos (Figure S2A-S2C) (Gierten et al., 2020), from 54 hpf, right before the onset of pdgfrb+ progenitors’ expression of acta2 (Figure 2A-2C). We found that at 3 dpf the number of acta2+ VSMCs on the CaDI of treated embryos is greatly reduced (Figure 4G and 4J); acta2+ VSMCs on BCA and PCS at 4 dpf are also reduced significantly (Figure 4G and 4K), suggesting that blood flow reduction after 54 hpf impaired VSMC differentiation on CoW arteries.

To determine whether blood flow is required for short-term VSMC maintenance after differentiation on CoW arteries, we treated embryos with nifedipine from 4 dpf, after VSMC differentiation on CoW arteries, and imaged at 5 dpf. The number of acta2+ VSMCs on the CaDI and PCS in treated embryos was comparable to untreated controls (Figure S2D-S2F), suggesting that blood flow was required for differentiation but not for short-term maintenance of CoW VSMCs.

Blood flow-regulated transcription factor klf2a modulates CoW artery muscularization

Our results suggest that CoW ECs might express a flow dependent transcriptional program that regulates spatiotemporal dynamics of VSMC differentiation on CoW arteries. Previous studies showed that expression pattern of transcription factor Klf2 in arterial ECs closely follow the pattern of flow induced WSS (Lee et al., 2006; Parmar et al., 2006). Consistently, higher Klf2 expression can be induced by unidirectional pulsatile flow (Dekker et al., 2002). Furthermore, klf2 is implicated in VSMC migration on arterial blood vessels during mouse development: loss of klf2 leads to aorta VSMC deficiency (Wu, Bohanan, Neumann, & Lingrel, 2008). Interestingly, klf2a expression in primitive veins of zebrafish trunk prevents VSMC association (Stratman et al., 2020). Hence, we tested the hypothesis that klf2a might be spatiotemporally regulated in CoW ECs and modulate VSMC differentiation. We first imaged Tg(klf2a:h2b-egfp, kdrl:hras-mcherry)ig11/s896, which labels nuclei with active klf2a signaling, and quantified the number of klf2a+ ECs in CoW arteries (Chi et al., 2008; Steed et al., 2016). To account for overall changes in EC number during development (Ulrich, Ma, Baker, & Torres-Vazquez, 2011), and length of each vessel, we also imaged Tg(fli1:nls-gfp, kdrl:hras-mcherry)y7/s896, which labels all EC nuclei (Figure S3A-S3I) (Roman et al., 2002). From 32 hpf to 54 hpf, the number of klf2a+ EC in CaDI was comparable to PCS (Figure 5A-5C), however, the number of klf2a+ ECs in CaDI was significantly higher than BCA and PCS from 3 dpf (Figure 5A and 5D-5E), when acta2+ VSMCs differentiated on CaDI but not BCA and PCS (Figure 2B-2C, 2F, and S1A-S1C). Pearson correlation comparing the number of klf2a+ ECs with the number of acta2+ VSMCs in CaDI, BCA and PCS at 3 dpf showed strong positive correlation with r = 0.974 (Figure S5J). These results suggest that the number of klf2a+ ECs in CoW arteries is spatially correlated with VSMC differentiation.

Blood flow regulated transcription factor klf2a is expressed in circle of Willis (CoW) arteries

(A) Confocal live images of Tg(klf2a:h2b-gfp, kdrl:ras-mcherry)ig11/s896 and scheme representation of endothelial cells (ECs) in CoW arteries in zebrafish brain at 32 hour post fertilization (hpf), 54 hpf, 3 day post fertilization (dpf), and 4 dpf. Green channel represents klf2a:h2b-gfp, Red channel represents kdrl:ras-mcherry, and Merge panel combines both channels. Arrows point to the CoW arteries with klf2a:h2b-gfp signal. Scale bar = 50 μm

(B-E) Number of klf2a+ ECs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 32 hpf (n=6, 3 independent experiments) (B), 54 hpf (n=6, 2 independent experiments) (C), 3 dpf (n=11, 3 independent experiments) (D), and 4 dpf (n=12, 3 independent experiments) (E), ordinary one-way ANOVA analyses with Tukey’s multiple comparisons, represented with mean ± SD, ** p≤0.01, **** p≤0.0001

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

To further define the role of klf2a in VSMC differentiation on CoW arteries, we knocked down klf2a in Tg(acta2:mcherry, kdrl:gfp)ca8/zn1 with 11 ng MO injection at one to two-cell stage and imaged at 3 to 4 dpf (Nicoli et al., 2010; Whitesell et al., 2014; Zhong et al., 2006). We validated the effect of klf2a MO knockdown with Tg(klf2a:h2b-egfp, kdrl:hras-mcherry)ig11/s896 (Figure S3G-S3I) (Chi et al., 2008; Steed et al., 2016). Compared to uninjected control, klf2a morphants have significantly less acta2+ VSMCs on CaDI at 3 dpf but a normal number by 4 dpf (Figure 6A-6D), suggesting that klf2a knockdown delayed VSMC differentiation on CaDI. Together these data suggest that endothelial klf2a is required for the onset of VSMC differentiation on CoW arteries.

klf2a promotes vascular smooth muscle cell (VSMC) differentiation on anterior circle of Willis (CoW) arteries

(A-B) Confocal live images of Tg(acta2:mcherry; kdrl:gfp)ca8/zn1 and scheme representation of vascular endothelium and VSMCs on CoW arteries at 3 day post fertilization (dpf) (A) and 4 dpf (B) in uninjected control embryos, embryos injected with 11 ng klf2a morpholino (MO). Red channel represents acta2:mcherry, Green channel represents kdrl:gfp, and Merge panel combines both channels. Arrows point to the CoW arteries with acta2:mcherry signal. Scale bar = 50 μm

(C) Number of acta2+ VSMCs per 100 μm vessel length on caudal division of internal carotid arteries (CaDI), basal communicating artery (BCA), and posterior communicating segments (PCS) at 3 dpf in uninjected control (n=8, 2 independent experiments) and embryos injected with 11 ng klf2a MO at one to two cell stage (n=8, 2 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD, ** p≤0.01

(D) Number of acta2+ VSMCs per 100 μm vessel length on CaDI, BCA, and PCS at 4 dpf in uninjected control (n=8, 3 independent experiments) and embryos injected with 11 ng klf2a MO at one to two cell stage (n=8, 3 independent experiments), two-tailed Mann–Whitney test on each vessel’s comparison, represented with mean ± SD

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, l-CaDI: left caudal division of internal carotid artery, r-CaDI: right caudal division of internal carotid artery, BCA: basal communicating artery, l-PCS: left posterior communicating segment, r-PCS: right posterior communicating segment

Discussion

In this study, we employed confocal live imaging of fluorescence transgenic zebrafish embryos to investigate the spatiotemporal dynamics of vascular smooth muscle cell (VSMC) differentiation on the cerebral vasculature (CoW), which comprises major arteries supplying blood to the vertebrate brain. Our observations revealed that CoW morphogenesis precedes the expression of arterial endothelial genes. Specifically, mural cell progenitors marked by pdgfrb+ initiate the expression of the VSMC marker acta2 subsequent to their recruitment to CoW arteries. Notably, VSMC differentiation occurs earlier in the anterior CoW arteries compared to their posterior counterparts, owing to the elevated wall shear stress (WSS) resulting from a higher velocity of red blood cells in the incoming blood flow. To investigate the regulatory role of blood flow on the spatiotemporal dynamics of VSMC differentiation, we employed an in vitro co-culture assay along with genetic manipulation and drug treatments. Our findings indicate that blood flow indeed governs the timing and location of VSMC differentiation on CoW arteries. Moreover, we observed that the flow-responsive transcription factor klf2a is activated in a gradient manner from anterior to posterior CoW arteries, preceding the onset of VSMC differentiation. Knockdown experiments targeting klf2a revealed a delay in VSMC differentiation specifically in the anterior CoW arteries. Taken together, these findings highlight the role of endothelial klf2a activation by blood flow as a mechanism that promotes VSMC differentiation on CoW arteries within the brain (see Figure 7 for a visual summary).

Schematic model of the developmental muscularization of the CoW arteries

The model shows how blood flow generate higher hemodynamics in anterior CoW arteries like the caudal division of internal carotid artery (CaDI) via the activation of endothelial klf2a signaling. Other posterior CoW arteries with straight shape like the posterior communicating segment (PCS) experience less hemodynamic force and showed moderate klf2a activation and VSMC differentiation.

Abbreviations: hpf: hour post fertilization, dpf: day post fertilization, EC: endothelial cell, VSMC: vascular smooth muscle cell, CaDI: caudal division of internal carotid artery, PCS: posterior communicating segment

Previous research shows that in zebrafish trunk, as soon as recruited to the dorsal aorta, VSMCs express acta2 and transgelin (tagln), suggesting simultaneous recruitment and differentiation from sclerotome progenitors (Ando et al., 2016; Stratman et al., 2017). On CoW arteries in zebrafish brain, however, VSMC differentiation occurs after pdgfrb+ progenitor recruitment and proceeds from anterior to posterior (Figure 2B-2F). The observed spatiotemporal dynamics of CoW VSMC differentiation again highlights organotypic development of blood vessels.

Our data suggests klf2a mediate blood flow regulation of VSMC differentiation on brain arteries, and thus raises the question of how klf2a transduces endothelial signals to mural cell progenitors and VSMCs. Notch signaling appears a plausible downstream effector of klf2a activation, as previous research suggests Notch responds to flow in heart valve development (Fontana et al., 2020), possibly downstream of klf2 (Duchemin, Vignes, & Vermot, 2019). Wnt signaling is another possible downstream effector of klf2a, as previous research suggests endocardial klf2 upregulates Wnt signaling in neighboring mesenchymal cells in heart valve development (Goddard et al., 2017). The roles of Notch and Wnt signaling in organotypic VSMC differentiation on brain arteries remain to be determined.

The expression of klf2a in CoW arterial ECs remained stable from 3 dpf to 4 dpf (Figure 5), which raises and interesting question on whether sustained klf2a expression supports further maturation of VSMCs, as acta2+ VSMCs on CaDI showed distinct morphology at 4 dpf compared with 3 dpf, when they started differentiation from pdgfrb+ mural cell progenitors (Figure 2C-2D and S1). Previous research found that acta2+ VSMCs on the BCA and PCS express pericyte enriched abcc9 (ATP binding cassette subfamily C member 9) at 4 dpf (Ando et al., 2022), when they started differentiation from pdgfrb+ mural cell progenitors and expressing acta2; abcc9 expression is gradually lost from 5 dpf to 6 dpf (Ando et al., 2022). In addition, acta2+ VSMCs on CaDI, BCA, and PCS still retain expression of pdgfrb at 6 dpf (Ando et al., 2021). Thus, it is possible that stable expression of klf2a in CoW arterial ECs supports further VSMC maturation, as indicated by expression of tagln (Colijn, Nambara, & Stratman, 2023). Another aspect of VSMC maturation is its function to confer vascular tone. A previous study found that VSMC covered vessels in zebrafish brain dilate as early as 4 dpf and constrict at 6 dpf (Bahrami & Childs, 2020). Future study may focus on the association between expression of different VSMC markers and VSMC functional maturation.

Another question is how flow activates klf2a in brain arteries. In a hypoxia-induced pulmonary hypertension model, klf2 is activated by G-protein coupled receptor mediated Apelin signaling (Chandra et al., 2011). In heart valve development, endocardial klf2a is upregulated by membrane bound mechanosensitive channels trpp2 and trpv4 (Heckel et al., 2015). The pathway through which flow activates klf2a in brain arterial ECs remains unknown. The CoW VSMC phenotype of klf2a morphants is similar to, but do not fully recapitulate the phenotypes of tnnt2a morphants or nifedipine treated embryos (Figure 4E-4L and Figure 6). A proximal explanation is compensation by paralogous klf2b in zebrafish. Further characterization of CoW VSMC development in klf2a and klf2b genetic mutants (Rasouli et al., 2018; Steed et al., 2016) may help determine whether klf2b compensates klf2a in CoW VSMC differentiation.

In addition to WSS, transmural pressure and associated mechanical stretch is another mechanical input of vascular wall that may contribute to VSMC differentiation. VSMCs autonomously sense and adapt to mechanical stretch (Haga, Li, & Chien, 2007). Cyclic stretching activates arterial VSMC production of certain components of extracellular matrix (ECM) (Leung, Glagov, & Mathews, 1976). In turn, specific interactions between ECM and integrins enable VSMCs to sense mechanical stretch (Goldschmidt, McLeod, & Taylor, 2001; Wilson, Sudhir, & Ives, 1995). Connection to ECM is essential for VSMC force generation (Milewicz et al., 2017). VSMCs express stretch activated trpp2, which enables myogenic response to autoregulate resting arterial diameter (Sharif-Naeini et al., 2009). Interestingly, ex vivo stretching promotes expression of contractile proteins, such as acta2 and tagln in VSMCs on murine portal veins (Albinsson, Nordstrom, & Hellstrand, 2004; Turczynska, Hellstrand, Sward, & Albinsson, 2013). In addition, excessive mechanical stretch promotes VSMC dedifferentiation and inflammation (Cao et al., 2017; Wang et al., 2018). Further investigation is needed to determine the potential role of increasing transmural pressure and associated mechanical stretch during development in VSMC differentiation.

There are a few limitations to our current study. Genetic manipulation and drug treatment that reduce heart rate would also reduce nutrients carried by blood flow, and the effects of nutrient and flow reduction could not be uncoupled in live zebrafish embryos. In addition, these methods are only capable of qualitative reduction of flow but not specific dampening of pulsations. In vitro three-dimensional (3D) vascular culture models, which combine ECs and mural cells (Mirabella et al., 2017; Vila Cuenca et al., 2021), could be further optimized to simulate complex geometry of brain arteries. Combining these models with microfluidics, which allows precise calibration of nutrient composition in culture media, flow velocity and pulse, would enable more thorough analysis of endothelial mechanotransduction and its contribution to VSMC differentiation (Abello, Raghavan, Yien, & Stratman, 2022; Gray & Stroka, 2017; Griffith et al., 2020).

We used nifedipine to temporally reduce heart rate and blood flow. Nifedipine is a blocker of L-type voltage-dependent calcium channels (VDCC) (Quevedo et al., 1998). A previous study shows that ML218, a T-type VDCC-selective inhibitor, tends to increase VSMC differentiation (Ando et al., 2022). In addition to the difference in drug targets, we also noted that in the previous study, the increase in VSMC differentiation only occur on anterior metencephalic central arteries (AMCtAs) that are more than 40 μm away from the BCA; these AMCtAs are much smaller than CoW arteries and have different geometry (Ando et al., 2022). Although the most obvious effect of nifedipine in zebrafish embryos is heart rate and blood flow reduction (Gierten et al., 2020), it is possible that nifedipine affect VSMC through VDCCs. Further investigation is needed to uncouple the roles of different VDCCs in VSMC differentiation and their association with hemodynamics.

We used Tg(flt4:yfp, kdrl:ras-mcherry)hu4881/s896 to show CoW arterial specification from primitive venous ECs (Chi et al., 2008; Hogan et al., 2009). kdrl is not the best arterial marker as its expression is only enriched but not restricted to arteries. While the use of the two fluorescence transgenic lines helped us establish the temporal sequence of CoW morphogenesis, arterial specification and VSMC differentiation, analysis of additional arterial and venous markers is needed to fully characterize arterial specification during vertebrate brain vascular development.

In summary, our work identifies flow induced endothelial klf2a activation as a mechanism that regulates spatiotemporal dynamics of VSMC differentiation on CoW arteries in the vertebrate brain. Our data may help advance approaches to regenerate differentiated VSMCs in cerebrovascular diseases. It would be important to further investigate mechanisms upstream and downstream of klf2a activation in brain arterial ECs and additional mechanotransduction pathways that may contribute to VSMC differentiation on brain arteries.

Material and Methods

Zebrafish husbandry and transgenic lines

Zebrafish were raised and maintained at 28.5 °C using standard methods. Protocols are approved by the Yale Institutional Animal Care and Use Committee (2020-11473). TgBAC(pdgfrb:egfp)ncv22 was generated by Ando et al., has been reported that the pdgfrb:egfp+ cells express other pericyte enriched marker abcc9 (ATP binding cassette subfamily C member 9) (Ando et al., 2016; Ando et al., 2019). By this transgenic line, they also observed that pdgfrb mutant lack brain pericytes and exhibit loss of vascular smooth muscle coverage (Ando et al., 2021). Thus, we used TgBAC(pdgfrb:egfp)ncv22 to study mural cell coverage of vascular endothelium in zebrafish CoW arteries. All transgenic lines in Table 1 were established previously.

Confocal fluorescence microscopy

Zebrafish embryos were raised in 0.003% 1-phenyl-2-thiourea (PTU, phenylthiocarbamide, or n-phenylthiourea, Sigma P7629) from gastrulation stage to prevent pigmentation. Embryos imaged live by confocal fluorescence microscopy were anesthetized in 0.1% tricaine methanesulfonate (TMS, MS-222, or Syncaine, Western Chemical, NC0872873) and mounted in 1% low melt agarose within glass bottom microwell dishes. Fluorescence images were captured with an upright Zeiss LSM 980 confocal microscope using a 20X objective.

Image analysis

Confocal fluorescence images were analyzed with Imaris microscopy image analysis software (Bitplane, Oxford Instruments). Average fluorescence intensity of mcherry in Tg(flt4:yfp, kdrl:hras-mcherry)hu4881/s896 was estimated by generating volume objects covering the artery of interest with Surface module. Vessel lengths of the arteries were manually traced with Filament module. Numbers of pdgfrb+ mural cell progenitors, acta2+ vascular smooth muscle cells, and klf2a+ and fli1+ endothelial nucleus were counted with Spots module.

Morpholino injections

Morpholino antisense oligonucleotides were synthesized by Gene Tools. Morpholinos in Table 2 were validated previously. Optimized dose injected into each embryo are listed. Uninjected siblings were used as controls.

Nifedipine treatment

Nifedipine was dissolved into 20 mM in dimethyl sulfoxide (DMSO). The stock solution was diluted into 25 μM in egg water with 0.003% PTU to treat zebrafish embryos from 54 hpf to 3 dpf or 4 dpf. The stock solution was diluted into 20 μM in egg water with 0.003% PTU to treat zebrafish embryos from 4 dpf to 5 dpf. The same volume of DMSO was added into egg water with 0.003% PTU for sibling control embryos.

Microangiography

Embryos were anesthetized in 0.1% tricaine methanesulfonate, placed on an agarose mold, and injected pericardially with 4 nL Tetramethylrhodamine Dextran (2,000,000 molecular weight, Thermo Fisher) at a concentration of 10 mg/mL. Subsequently, the vasculature of embryos was checked for Dextran fluorescence signal under a stereomicroscope (Olympus, MVX10). Embryos with Dextran fluorescence were mounted in 1% low melt agarose within glass bottom microwell dishes and imaged with confocal microscope.

Red blood cell velocity

Red blood cell (RBC) velocity was measured by used axial line scanning to analyze zebrafish in the Tg(kdrl:gfp;gata1:DsRed)zn1/sd2 background, in which blood vessels and RBCs are labeled with green and red fluorescent protein, respectively (Cross et al., 2003; Traver et al., 2003). As described previously (Barak et al., 2021), this axial line scan was conducted in a Zeiss LSM980 confocal microscope. A line was drawn inside the lumen of CoW vessels and parallel to the longitudinal axis of the vessel. The average length of the line was 21.21 μm. The scan speed was chosen as maximum per second for 20,000 cycles (total scan duration was 5.86 s; line scan time is 292.80 μs). A kymograph was exported in TIFF format after every measurement by ZEN 3.5 and analyzed RBC recognition and segmentation by using Fiji (ImageJ). Firstly, the RBC kymography image was processed using ‘noise despeckle’ and ‘enhance contrast’ steps. The brightness and contrast were appropriately adjusted if needed. Next, the RBCs were segmented, thresholding was adjusted, and the slope of the RBCs was calculated automatically using the ‘Analyze particle’ option. Finally, the velocity of each RBC was calculated using the following equation: velocity = (1/tan(a)) * (length of the y axis of the kymograph/duration of the total scan). We averaged systolic peak velocities and end diastolic velocities to measure the mean RBC velocity in the CoW vessels.

Computational fluid dynamic (CFD) simulation

Computational fluid dynamic (CFD) simulation was performed according to the geometry model of the circle of Willis (CoW) at 54 hpf, 3 and 4 dpf as previously described with modification (ANSYS, 2014; Barak et al., 2021). First, three-dimensional (3D) geometry models of CoW arteries were reconstructed from confocal z-stack microangiography images of Dextran injected Tg(kdrl:gfp) embryos with the Filament module in Imaris (9.9.0, Bitplane, Oxford instruments). These CoW 3D geometry models were pre-processed in ANSYS SpaceClaim 2022 R1 software. For example, the computational mesh on the CoW vascular wall was created and the boundary conditions were set up for the simulation. Meshing orthogonal quality was calculated, and objects less than 0.01 were excluded. Inlets and outlets were determined in the 3D model for the flow simulation. Next, the pre-processed models were imported in ANSYS Fluent 2022 R1 software. The non-Newtonian behavior of blood flow was modeled using the Carreau model which was provided by ANSYS. Blood density was considered as 1,060 kg/m3. The computational simulation used average RBC velocity among the CoW arteries at 54 hpf (532.72 μm/s), 3 dpf (602.34 μm/s) and 4 dpf (552.83 μm/s), respectively. After the CoW geometry and numerical solution of blood flow were ready, the flow simulation was implemented with 200 iterations to identify the magnitude and distribution of the WSS in the CoW.

Cell Culture

Pooled primary Human Umbilical Vein Endothelial Cells (HUVEC) (PCS-100-010TM, ATCC) were seeded at an initial concertation of 5,000 cells per cm2, in 1x M119 media (110343-023, Gibco) supplemented with 16% FBS (Gibco), 84 µg/mL of heparin sodium salt (H3393, Sigma-Aldrich), 25 µg/mL of endothelial cell growth supplement (02-102, EMD Millipore Corp.) and 1x Antibiotic-antimycotic solution (15240-062, Gibco). Cell cultures were maintained at 37°C, 5% CO2 and 95% humidity, until the cell reached 80% confluence.

To facilitate cell visualization in cell co-cultures, Human Brain Vascular Pericytes (HBVP) (1200, ScienCell) were transfected using lentiviral particles (LVP310, GenTarget, Inc.) to induce GFP expression under EF1a promotor. Cells cultures were initiated by seeding 5,000 cells per cm2 in 175 cm2 plastic flasks pre-coated with gelatin and 1x DMEM (11995-065, Gibco) supplemented with 10% FBS (Gibco), and 1x Antibiotic-antimycotic solution under the same cell culture conditions described above. When cell cultures were at 80% confluence, 200 µL containing 2 x 106 GFP-lentiviral particles were added to each 175 flasks. After 72 hours, fresh cell culture media supplemented with 10µg/mL of Blasticidin (15205, Sigma-Aldrich) were added to each flask to select the positive transfected cells.

Flow assays and immunostaining

HBVP cells were harvested and seeded at a concentration of 1.3 x 105 cells in 0.4 optical plastic flow microslides (80176, Ibidi) precoated with 1 mg/mL gelatin and incubated for 24 hours under standard culture conditions. After the initial incubation, 100 µg/mL of collagen I (354249, Corning) diluted in DMEM cell culture media was added to the slides to create a thin layer on top of the HBVP cells. After 2 hours, the media was removed and 2.5 x 105 HUVECs were seeded on top of the collagen I layer and incubated for additional 24 hours. After cell co-cultures were established, the slides were exposed to laminar (15 dyn/cm2) or pulsatile (12-15 dyn/cm2) flow for 24 hours, implementing a peristaltic pump adapted to produce different types of flow (Abello, Raghavan, Yien, & Stratman, 2022). After 24 hours, cultures were rinsed with 1x PBS and fixed for 30 minutes in 4% paraformaldehyde at room temperature. Cell cultures were immunostained with α-Smooth muscle actin D4K9N XP® rabbit monoclonal antibody (19245, Cell Signaling), followed by Alexa Fluor 633 goat anti-rabbit IgG (A21071, Invitrogen). Confocal images were obtained using a 40x objective with a W1 Spinning Disk confocal microscope, a Fusion camera, and the Nikon Eclipse Ti2-E base. Fiji image processing software was used for image analysis and fluorescence intensity quantification.

Statistics

All statistical analyses were performed with GraphPad Prism (version 10.0.3). Mann–Whitney test was used to compare two groups to test mean differences (protein level, morpholino and nifedipine treatment). Two-way ANOVA followed by Tukey’s multiple comparisons was used to test mean differences compared more than two groups under different conditions (average fluorescence intensity, cell number, vascular diameter). Ordinary one-way ANOVA with Tukey’s multiple comparisons was used to test mean differences among three groups (RBC velocity, WSS, and cell number). Pearson correlation analysis was used to compare the strength and direction of a linear relationship between two variables with a covariance (WSS vs acta2+ EC, acta2+ EC vs klf2+ EC).